---
title: "Stage 1: Data Loading, QC & Normalisation"
subtitle: "Maris Piper Phosphate Stress — E-MTAB-2990"
author: "MSc Bioinformatics Student"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

---

# Background: What Are We Doing?

## The Experiment in Plain English

A scientist at the James Hutton Institute (Scotland) grew **4 varieties of potato** in water tanks — some tanks had **normal phosphate** (Reg P), others had **very low phosphate** (Low P). He then took root samples, extracted RNA, and ran them on a **microarray chip** that measures the activity of ~42,000 genes at once.

The result: **24 raw data files** — one per plant sample.

**Our job:** Re-analyse those 24 files using modern tools to find which genes in **Maris Piper** respond to low phosphate.

## What Does Our Pipeline Do?

```
24 raw .txt files
        ↓
   Load with limma        ← Step 1: Read all files into R
        ↓
   Quality Control        ← Step 2: Check if arrays look good
        ↓
 Background Correction    ← Step 3: Remove noise
        ↓
    Normalisation         ← Step 4: Make arrays comparable
        ↓
  Filter bad probes       ← Step 5: Remove unreliable signals
        ↓
     PCA plot             ← Step 6: Check samples cluster correctly
        ↓
  Save clean data         ← Step 7: Ready for differential expression
```

---

# Load Packages

```{r load-packages}
# ── Why BiocManager for limma? ────────────────────────────────────────────
# limma is a Bioconductor package — it is NOT on CRAN.
# install.packages("limma") will FAIL — you must use BiocManager.
# Reference: Ritchie et al. (2015) Nucleic Acids Research 43(7):e47
# ─────────────────────────────────────────────────────────────────────────

# Install BiocManager if not present
if (!require("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

# Install limma via Bioconductor (only runs if not already installed)
if (!require("limma", quietly = TRUE)) {
  BiocManager::install("limma", ask = FALSE)
}

library(limma)        # Microarray analysis — Ritchie et al. (2015)
library(ggplot2)      # Plotting — Wickham (2016)
library(RColorBrewer) # Colour palettes
library(pheatmap)     # Heatmaps (needed later in Stage 2)

cat("Packages loaded:\n")
cat("  limma version:", as.character(packageVersion("limma")), "\n")
cat("  ggplot2 version:", as.character(packageVersion("ggplot2")), "\n")
```

---

# Define File Paths

```{r set-paths}
# ── Why define paths at the top? ─────────────────────────────────────────
# Centralising paths means one edit updates the whole notebook
# ─────────────────────────────────────────────────────────────────────────

data_path     <- "C:/Users/hrush/Desktop/UoS/project/Maris piper/data/raw/ZIP/"
metadata_path <- "C:/Users/hrush/Desktop/UoS/project/Maris piper/metadata/"
results_path  <- "C:/Users/hrush/Desktop/UoS/project/Maris piper/results/"

# Create results folder if it doesn't exist
if (!dir.exists(results_path)) {
  dir.create(results_path, recursive = TRUE)
  cat("Created results folder\n")
} else {
  cat("Results folder exists ✓\n")
}
```

---

# Read SDRF Metadata

```{r read-sdrf}
# ── What is the SDRF? ─────────────────────────────────────────────────────
# Sample and Data Relationship Format — links filenames to metadata
# check.names = FALSE keeps column names like "Characteristics[cultivar]"
# intact rather than converting them to Characteristics.cultivar.
# ─────────────────────────────────────────────────────────────────────────

sdrf <- read.delim(
  file        = paste0(metadata_path, "E-MTAB-2990.sdrf.txt"),
  check.names = FALSE
)

cat("SDRF dimensions:", nrow(sdrf), "rows x", ncol(sdrf), "columns\n")
cat("\nKey columns:\n")
print(colnames(sdrf)[c(3, 5, 26, 29)])
```

---

# Build Targets Table

```{r build-targets}
# ── Why only these 3 columns? ─────────────────────────────────────────────
# The SDRF has 29 columns but most are administrative metadata
# (lab names, protocol references, scanner info).
# We only need: which file, which cultivar, which condition.
# ─────────────────────────────────────────────────────────────────────────

targets <- data.frame(
  FileName  = sdrf[["Array Data File"]],
  Cultivar  = sdrf[["Characteristics[cultivar]"]],
  Condition = sdrf[["Factor Value[growth condition]"]],
  stringsAsFactors = FALSE
)

cat("=== SAMPLE COUNT PER CULTIVAR × CONDITION ===\n")
print(table(targets$Cultivar, targets$Condition))

cat("\n=== MARIS PIPER FILES ===\n")
print(targets[targets$Cultivar == "Maris Piper", ])
```

> **Is 6 samples enough?** Yes — limma's empirical Bayes was specifically designed for small n. Many published microarray studies use 3 replicates per group. Reference: Smyth (2004) *Stat Appl Genet Mol Biol* 3(1):Article3.

> **What is 12601?** A potato breeding line referred to by number rather than a commercial name — it was never released commercially. It serves as a low phosphate-use-efficiency comparison cultivar in this experiment.

---

# Load Raw Agilent Data

```{r load-raw-data}
# ── read.maimages() for Agilent one-colour arrays ─────────────────────────
# source = "agilent"   → reads Agilent Feature Extraction .txt format
# green.only = TRUE    → one-colour array (Cy3 dye only, no Cy5)
# other.columns        → also load the quality flag gIsWellAboveRG
#                        (1 = probe signal above background, 0 = noise)
# ─────────────────────────────────────────────────────────────────────────

cat("Loading 24 Agilent files — this may take 1-2 minutes...\n")

raw <- read.maimages(
  files         = targets$FileName,
  source        = "agilent",
  path          = data_path,
  green.only    = TRUE,
  other.columns = "gIsWellAboveRG"
)

cat("\nRaw data loaded successfully!\n")
cat("Matrix dimensions:", nrow(raw$E), "probes x", ncol(raw$E), "samples\n")
cat("\nFirst 5 rows of raw intensity matrix (first 3 samples):\n")
print(round(raw$E[1:5, 1:3], 2))
```

> **Explaining the numbers:** Probe 1 shows ~1474–2748 across 3 samples — a highly expressed, consistent gene. Probe 2 shows ~48–57 — borderline noise. Probe 5 shows ~2461–4300 — highly expressed and variable. After filtering, the noise probes (like probe 2) will be removed.

---

# Quality Control — Raw Data

```{r qc-raw-boxplot, fig.width=14, fig.height=6}
# ── Why log2 for plotting? ────────────────────────────────────────────────
# Raw intensities range from ~10 to ~100,000 — too wide to visualise
# log2 compression makes differences visible
# Also: log2(fold change) = 1 means 2-fold change — standard unit
# ─────────────────────────────────────────────────────────────────────────

cultivar_colours <- c(
  "Maris Piper"  = "#E41A1C",
  "Stirling"     = "#377EB8",
  "Pentland Dell"= "#4DAF4A",
  "12601"        = "#FF7F00"
)
col_vec <- cultivar_colours[targets$Cultivar]

boxplot(
  log2(raw$E),
  las      = 2,
  col      = col_vec,
  main     = "Raw log2 Intensities — Pre-Normalisation",
  ylab     = "log2(Intensity)",
  names    = paste0(substr(targets$Cultivar, 1, 2), "_",
                    targets$Condition, "_",
                    rep(1:3, 8)),
  cex.axis = 0.7
)
legend("topright", legend = names(cultivar_colours),
       fill = cultivar_colours, cex = 0.8, title = "Cultivar")
abline(h = median(log2(raw$E), na.rm = TRUE),
       col = "black", lty = 2, lwd = 1.5)
```

```{r qc-raw-density, fig.width=10, fig.height=6}
plotDensities(log2(raw$E), col = col_vec,
              main = "Signal Density — Pre-Normalisation",
              legend = FALSE)
legend("topright", legend = names(cultivar_colours),
       col = cultivar_colours, lty = 1, lwd = 2, cex = 0.8)
```

---

# Background Correction

```{r background-correction}
# ── normexp method ────────────────────────────────────────────────────────
# Models background as normally distributed, signal as exponential
# Mathematically prevents negative values (unlike simple subtraction)
# offset = 16 adds a small constant to stabilise near-zero values
# Reference: Ritchie et al. (2007) Nucleic Acids Research 35(19):e118
# ─────────────────────────────────────────────────────────────────────────

bc <- backgroundCorrect(raw, method = "normexp", offset = 16)

cat("Background correction complete\n")
cat("Any negative values?", any(bc$E < 0, na.rm = TRUE), "\n")
cat("Min value after correction:", round(min(bc$E, na.rm = TRUE), 3), "\n")
```

---

# Quantile Normalisation

```{r normalise}
# ── Why quantile normalisation? ───────────────────────────────────────────
# Makes all 24 array distributions identical by replacing probe ranks
# with the cross-array average at that rank
# Removes technical differences while preserving biological signal
# Assumption: most genes (~95%+) are NOT differentially expressed — 
# valid here since we compare 2 conditions in the same tissue type
# Reference: Bolstad et al. (2003) Bioinformatics 19(2):185-193
# ─────────────────────────────────────────────────────────────────────────

norm <- normalizeBetweenArrays(bc, method = "quantile")

cat("Quantile normalisation complete\n")
cat("All 24 arrays now have identical distributions\n")

# Why was Ma_RegP_1 higher before? It was slightly over-labelled or
# over-scanned — a technical artefact. Normalisation corrects this.
# The original paper used GeneSpring v7.3 (per-chip median normalisation)
# Quantile normalisation is statistically superior (Bolstad et al. 2003)
```

```{r qc-norm-boxplot, fig.width=14, fig.height=6}
boxplot(
  norm$E,
  las      = 2,
  col      = col_vec,
  main     = "Normalised Intensities — Post-Normalisation",
  ylab     = "log2(Normalised Intensity)",
  names    = paste0(substr(targets$Cultivar, 1, 2), "_",
                    targets$Condition, "_",
                    rep(1:3, 8)),
  cex.axis = 0.7
)
legend("topright", legend = names(cultivar_colours),
       fill = cultivar_colours, cex = 0.8, title = "Cultivar")
abline(h = median(norm$E, na.rm = TRUE),
       col = "black", lty = 2, lwd = 1.5)
```


---

# Evidence-Based Probe Filtering

```{r filter-probes}
# ── Why use negative controls as the threshold? ───────────────────────────
# Agilent arrays contain built-in negative control probes (ControlType=-1)
# These sequences are designed NOT to match any biological sequence
# Their intensity = pure background noise with no real signal
# The 95th percentile of their intensities = the noise ceiling
# Any probe at or below this level is indistinguishable from noise
# The 1.1× safety margin ensures we are clearly ABOVE noise, not just at it
# Reference: Gentleman et al. (2004); Shi et al. (2006) MAQC
# ─────────────────────────────────────────────────────────────────────────

# First: apply a broad pre-filter to create norm_filtered
# We keep probes with log2 > 5 in >= 3 samples as an initial step
# Then apply the evidence-based negative control filter on top
pre_expressed <- rowSums(norm$E > 5) >= 3
norm_filtered <- norm[pre_expressed, ]

cat("Pre-filter (log2 > 5, n>=3): retained", nrow(norm_filtered$E), "probes\n\n")

# Now apply evidence-based filter using negative control probes
neg_controls <- norm_filtered$E[norm_filtered$genes$ControlType == -1, ]
cat("Number of negative control probes:", nrow(neg_controls), "\n")

# Per-array 95th percentile
neg95 <- apply(neg_controls, 2, function(x) quantile(x, 0.95))
cat("95th percentile noise ceiling per array:\n")
print(round(neg95, 3))

# Per-array cutoff matrix with 1.1× safety margin
cutoff <- matrix(1.1 * neg95,
                 nrow = nrow(norm_filtered$E),
                 ncol = ncol(norm_filtered$E),
                 byrow = TRUE)

# Keep probes above cutoff in at least 3 samples
expressed <- rowSums(norm_filtered$E > cutoff) >= 3
norm_filtered2 <- norm_filtered[expressed, ]

cat("\nProbes before evidence-based filter:", nrow(norm_filtered$E), "\n")
cat("Probes after evidence-based filter: ", nrow(norm_filtered2$E), "\n")
cat("Probes removed:                     ",
    nrow(norm_filtered$E) - nrow(norm_filtered2$E), "\n")
cat("% retained:                         ",
    round(nrow(norm_filtered2$E)/nrow(norm_filtered$E)*100, 1), "%\n")
cat("\nTarget range: 50-70% (Shi et al. 2006 MAQC recommendation)\n")

# Use the evidence-based filtered object going forward
norm_filtered <- norm_filtered2
```

```{r threshold-comparison}
# ── Show what different thresholds would have retained ────────────────────
# This justifies why our evidence-based approach is superior to arbitrary
# thresholds — the data itself tells us where the noise floor is
# ─────────────────────────────────────────────────────────────────────────

cat("=== THRESHOLD COMPARISON ===\n")
thresholds <- c(5, 5.5, 6, 6.5, 7, 7.5, 8)
for (t in thresholds) {
  n_pass <- sum(rowSums(norm$E > t) >= 3)
  pct    <- round(n_pass / nrow(norm$E) * 100, 1)
  flag   <- if (pct > 70) "← too permissive" else
            if (pct < 50) "← too strict" else "✓ acceptable range"
  cat("Threshold", t, "→", n_pass, "probes (", pct, "%)", flag, "\n")
}
cat("\nEvidence-based (neg ctrl 95th pct × 1.1) →",
    nrow(norm_filtered$E), "probes (",
    round(nrow(norm_filtered$E)/nrow(norm$E)*100, 1), "%) ✓ data-driven\n")
```

---

# ── CRITICAL FIX: Assign Row Names ────────────────────────────────────────

```{r assign-rownames}
# ── WHY THIS STEP IS ESSENTIAL ────────────────────────────────────────────
# read.maimages() does NOT automatically assign row names to the expression
# matrix. It tracks probes by integer position internally.
# topTable() in Stage 2 returns results with those integer positions as
# row names (e.g., "26832", "35566").
# If norm_filtered$E has NULL rownames, then mp_expr will also have NULL
# rownames, and intersect() in the heatmap code will find 0 matches.
#
# FIX: Assign sequential integer positions as character row names NOW
# so that Stage 2 can match them correctly.
# ─────────────────────────────────────────────────────────────────────────

# Assign row names as character strings of row positions
rownames(norm_filtered$E) <- as.character(seq_len(nrow(norm_filtered$E)))

# Also assign to the genes annotation table if present
if (!is.null(norm_filtered$genes)) {
  rownames(norm_filtered$genes) <- as.character(seq_len(nrow(norm_filtered$genes)))
}

cat("=== ROWNAME FIX APPLIED ===\n")
cat("norm_filtered$E rownames are NULL:", is.null(rownames(norm_filtered$E)), "\n")
cat("First 5 rownames:", head(rownames(norm_filtered$E), 5), "\n")
cat("Last 5 rownames: ", tail(rownames(norm_filtered$E), 5), "\n")
cat("\nThis ensures Stage 2 heatmaps work correctly.\n")
```

---

# PCA — All 24 Samples

```{r pca-all-samples, fig.width=9, fig.height=7}
# ── PCA explained ─────────────────────────────────────────────────────────
# Reduces 38,885 dimensions to 2 for visualisation
# scale. = TRUE standardises each probe to unit variance
# Samples close together = similar gene expression profiles
# ─────────────────────────────────────────────────────────────────────────

pca_all <- prcomp(t(norm_filtered$E), scale. = TRUE)
var_exp <- summary(pca_all)$importance[2, ] * 100

pca_df <- data.frame(
  PC1       = pca_all$x[, 1],
  PC2       = pca_all$x[, 2],
  Cultivar  = targets$Cultivar,
  Condition = targets$Condition
)

ggplot(pca_df, aes(x = PC1, y = PC2,
                    colour = Cultivar,
                    shape  = Condition)) +
  geom_point(size = 5, alpha = 0.85) +
  stat_ellipse(aes(group = Cultivar), type = "norm",
               linetype = 2, alpha = 0.4) +
  scale_colour_manual(values = cultivar_colours) +
  scale_shape_manual(values = c("Low P" = 17, "Reg P" = 16)) +
  labs(
    title    = "PCA — All 24 Samples Post-Normalisation",
    subtitle = "Shapes: ▲ = Low P   ● = Reg P",
    x        = paste0("PC1 (", round(var_exp[1], 1), "% variance)"),
    y        = paste0("PC2 (", round(var_exp[2], 1), "% variance)")
  ) +
  theme_bw(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```

## PCA Scores Table — All Samples

```{r pca-table-all}
pca_table_all <- data.frame(
  Sample    = colnames(norm_filtered$E),
  Cultivar  = targets$Cultivar,
  Condition = targets$Condition,
  PC1       = round(pca_all$x[, 1], 2),
  PC2       = round(pca_all$x[, 2], 2),
  PC3       = round(pca_all$x[, 3], 2)
)

cat("=== PCA SCORES — ALL 24 SAMPLES ===\n")
print(pca_table_all)

cat("\n=== VARIANCE EXPLAINED ===\n")
print(data.frame(
  PC            = paste0("PC", 1:6),
  Variance_pct  = round(var_exp[1:6], 2),
  Cumulative_pct = round(cumsum(var_exp)[1:6], 2)
))
```

---

# PCA — Maris Piper Only

```{r pca-maris-piper, fig.width=9, fig.height=7}
# ── Why a separate MP PCA? ────────────────────────────────────────────────
# The all-24-sample PCA is dominated by cultivar differences
# A Maris Piper-only PCA shows whether Low P vs Reg P replicates
# cluster correctly within this cultivar — and reveals the outlier
# ─────────────────────────────────────────────────────────────────────────

mp_idx  <- which(targets$Cultivar == "Maris Piper")
pca_mp  <- prcomp(t(norm_filtered$E[, mp_idx]), scale. = TRUE)
var_mp  <- summary(pca_mp)$importance[2, ] * 100

# Label samples clearly
mp_labels <- paste0(
  ifelse(targets$Condition[mp_idx] == "Low P", "LowP", "RegP"),
  "_rep",
  ave(targets$Condition[mp_idx],
      targets$Condition[mp_idx],
      FUN = seq_along)
)

pca_mp_df <- data.frame(
  PC1       = pca_mp$x[, 1],
  PC2       = pca_mp$x[, 2],
  Condition = targets$Condition[mp_idx],
  Label     = mp_labels
)

# Identify outlier for highlighting
OUTLIER_FILE <- "US10310381_253303310046_S01_GE1_107_Sep09_1_4.txt"
outlier_pos  <- which(targets$FileName[mp_idx] == OUTLIER_FILE)
pca_mp_df$IsOutlier <- seq_len(nrow(pca_mp_df)) == outlier_pos

ggplot(pca_mp_df, aes(x = PC1, y = PC2,
                       colour = Condition,
                       shape  = IsOutlier,
                       label  = Label)) +
  geom_point(size = 6, alpha = 0.9) +
  geom_text(vjust = -1.2, size = 3.5, colour = "black") +
  scale_colour_manual(values = c("Low P" = "#E41A1C", "Reg P" = "#377EB8")) +
  scale_shape_manual(values = c("FALSE" = 16, "TRUE" = 17),
                     labels = c("Regular", "Outlier")) +
  labs(
    title    = "PCA — Maris Piper Only (6 Samples)",
    subtitle = "Triangle (▲) = identified outlier Reg P replicate",
    x        = paste0("PC1 (", round(var_mp[1], 1), "% variance)"),
    y        = paste0("PC2 (", round(var_mp[2], 1), "% variance)"),
    colour   = "Condition",
    shape    = "Sample type"
  ) +
  theme_bw(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```

## PCA Scores Table — Maris Piper

```{r pca-table-mp}
pca_mp_table <- data.frame(
  Label     = mp_labels,
  FileName  = targets$FileName[mp_idx],
  Condition = targets$Condition[mp_idx],
  PC1       = round(pca_mp$x[, 1], 2),
  PC2       = round(pca_mp$x[, 2], 2),
  IsOutlier = ifelse(seq_len(length(mp_idx)) == outlier_pos, "YES ⚠", "no")
)

cat("=== PCA SCORES — MARIS PIPER (6 SAMPLES) ===\n")
print(pca_mp_table)

cat("\n=== MARIS PIPER VARIANCE EXPLAINED ===\n")
print(data.frame(
  PC            = paste0("PC", 1:4),
  Variance_pct  = round(var_mp[1:4], 2),
  Cumulative_pct = round(cumsum(var_mp)[1:4], 2)
))
```

## Inter-Array Correlation Table

```{r correlation-table}
# ── Quantitative justification for the outlier decision ───────────────────
# Threshold: r > 0.90 for acceptable replicates
# Reference: Shi et al. (2006) MAQC, Nature Biotechnology 24:1151
# ─────────────────────────────────────────────────────────────────────────

mp_mat <- norm_filtered$E[, mp_idx]
colnames(mp_mat) <- mp_labels
cor_mat <- round(cor(mp_mat, method = "pearson"), 4)

cat("=== INTER-ARRAY PEARSON CORRELATION (Maris Piper) ===\n")
cat("Threshold: r > 0.90 (Shi et al. 2006, MAQC)\n\n")
print(cor_mat)

# Mean and min correlation per sample
cat("\n=== MEAN AND MIN CORRELATION PER SAMPLE ===\n")
mean_cor_df <- data.frame(
  Sample    = colnames(mp_mat),
  Condition = targets$Condition[mp_idx],
  Mean_r    = round(apply(cor_mat, 1, function(x) mean(x[x != 1])), 4),
  Min_r     = round(apply(cor_mat, 1, function(x) min(x[x != 1])), 4)
)
mean_cor_df$Decision <- ifelse(mean_cor_df$Min_r < 0.90,
                                "⚠ Below threshold",
                                "✓ Acceptable")
print(mean_cor_df)

# Save correlation table
write.csv(as.data.frame(cor_mat),
          paste0(results_path, "MP_interarray_correlations.csv"),
          row.names = TRUE)
```

---

# Save Results

```{r save-results}
# ── What is saved and why ─────────────────────────────────────────────────
# norm_filtered: the clean, filtered EList object with:
#   - $E       = expression matrix (38,885 × 24) WITH row names
#   - $genes   = probe annotation table WITH row names
#   - $targets = sample metadata
# targets: the cultivar/condition lookup table
# ─────────────────────────────────────────────────────────────────────────

save(
  norm_filtered,
  targets,
  file = paste0(results_path, "normalised_data.RData")
)

cat("=== SAVED: normalised_data.RData ===\n")
cat("Contents:\n")
cat("  norm_filtered$E dimensions:",
    nrow(norm_filtered$E), "probes x", ncol(norm_filtered$E), "samples\n")
cat("  norm_filtered$E rownames set:", !is.null(rownames(norm_filtered$E)), "\n")
cat("  First 3 rownames:", head(rownames(norm_filtered$E), 3), "\n")
cat("  targets rows:", nrow(targets), "\n")
cat("\nStage 1 complete. Proceed to Stage 2.\n")
```

---

# Session Info

```{r session-info}
sessionInfo()
```

---

# Summary of Stage 1

| Step | What we did | Evidence/Reference |
|---|---|---|
| Load packages | `BiocManager::install("limma")` | Ritchie et al. 2015, NAR |
| Read SDRF | Linked filenames to cultivar + condition | MAGE-TAB standard |
| Load raw data | `read.maimages()` Agilent one-colour | limma User's Guide |
| Background correction | `normexp`, offset=16 | Ritchie et al. 2007, NAR |
| Quantile normalisation | `normalizeBetweenArrays()` | Bolstad et al. 2003, Bioinformatics |
| Evidence-based filtering | Negative control 95th pct × 1.1 | Gentleman et al. 2004; Shi et al. 2006 MAQC |
| Row name assignment | `as.character(seq_len())` | Prevents Stage 2 crash |
| PCA | `prcomp()` | Standard dimensionality reduction |
| Outlier evidence | Inter-array correlation table | Shi et al. 2006 MAQC (r > 0.90) |
| Save | `normalised_data.RData` | With rownames intact |

**Next: Stage 2 — Differential Expression Analysis**
