---
title: "03 — Functional Enrichment Analysis"
subtitle: "Maris Piper Phosphate Stress | E-MTAB-2990"
author: "MSc Bioinformatics Student"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 7)
```

> **Stage 3 at a glance**
>
> - **Needs:** `results/DEG_results.RData` (Stage 2), `results/normalised_data.RData` (Stage 1), the raw Agilent files in `data/raw/` (re-read for probe annotation) and an **internet connection** to Ensembl Plants BioMart.
> - **Produces:** `results/Stage3_complete.RData` and `results/Key_genes_annotated.csv` (both loaded by Stage 4), GO enrichment tables, PGSC ID lists, annotation tables and figures (`.png`).
> - **Run time:** about 3-10 minutes, almost all of it the BioMart download (depends on Ensembl server load).
> - Terms used here (GO, KEGG, enrichment, background/universe, PGSC IDs, DMT vs DMG) are explained in the README **Glossary**.
> - The KEGG attempts described below were done outside this notebook; there is no KEGG code here.

---

# What Is This Stage Doing — And Why?

## The problem with a list of gene IDs

After Stage 2, we have two lists of differentially expressed genes (DEGs):

- **Analysis A** — Low Phosphate vs Regular Phosphate, with all samples
- **Analysis B** — Low Phosphate vs Regular Phosphate, outlier Reg P replicate removed

(The DEG counts are printed when the Stage 2 results are loaded below.)

These are just lists of identifiers like `PGSC0003DMT400069516`. On their own they mean nothing. Stage 3 asks: **what do these genes actually do?**

## What is functional enrichment?

Imagine you have a bag of a few hundred LEGO bricks, and you notice that 40 of them are wheel-shaped. You'd ask — is 40 wheels unusual? If wheels make up only 2% of all LEGO pieces but 14% of your bag, something is going on. That's enrichment.

We do the same thing with biology:

- We have a reference list of **all genes that were actually tested** — the probes that passed Stage 1 filtering, with Agilent control probes removed (the background, or "universe")
- We have our **DEG list** (the bag we're examining)
- We test: are certain biological functions over-represented in our DEG list?

The test used is a **hypergeometric test** (similar to Fisher's exact test). The result tells us which biological processes, molecular functions, or cellular components appear more than expected by chance.

## What is Gene Ontology (GO)?

Gene Ontology (GO) is a standardised vocabulary used across all of biology to describe what genes do. Every GO term has a unique ID (like `GO:0006796`) and falls into one of three categories:

| Category | What it describes | Example |
|---|---|---|
| Biological Process (BP) | What the gene participates in | "phosphate ion transport" |
| Molecular Function (MF) | What the gene does biochemically | "acid phosphatase activity" |
| Cellular Component (CC) | Where the gene product is located | "plasma membrane" |

## What is KEGG?

KEGG (Kyoto Encyclopedia of Genes and Genomes) organises genes into metabolic and signalling **pathway diagrams** — for example, the glycolysis pathway, or the plant hormone signalling pathway. KEGG enrichment tells you which pathways your DEGs are concentrated in.

## Why was this analysis challenging for potato?

Potato (*Solanum tuberosum*) is a non-model organism. Unlike Arabidopsis (which has been studied for decades), potato genome annotation is incomplete:

- The PGSC v3.4 assembly (2011) covers only ~60% of genes with functional information (author's estimate from the literature — *to verify*; the coverage actually achieved for our DEGs is computed below)
- The PGSC transcript IDs we have (`DMT` format) do not match KEGG's internal potato database
- No local Bioconductor annotation package (`org.St.eg.db`) is available

This is **normal and expected** for non-model organisms. We adapted our strategy accordingly and report exactly what worked.

## What strategy did we use?

We used three approaches in order:

| Approach | Tool | Outcome |
|---|---|---|
| KEGG REST API (direct) | httr → KEGG `/conv` endpoint | ❌ PGSC IDs not recognised by KEGG |
| GO enrichment via Ensembl Plants BioMart | biomaRt + clusterProfiler `enricher()` | ✅ Worked — nominal terms (counts computed below) |
| Key gene annotation (BioMart description lookup) | biomaRt getBM | ✅ Worked — named all key gene families |

KEGG via Arabidopsis orthologs was also attempted (151 orthologs mapped, according to the author's notes — done outside this notebook, so *to verify after re-run*) but returned no significant pathways. All negative results are reported transparently.

---

# Load Packages

```{r packages}
# ── What these packages do ────────────────────────────────────────────────
# clusterProfiler : runs GO enrichment statistics
# biomaRt         : connects to Ensembl Plants database to get annotations
# ggplot2         : makes plots
# dplyr           : data manipulation (filter, arrange, summarise)
# ggVennDiagram   : draws Venn diagrams
# ─────────────────────────────────────────────────────────────────────────

# Packages are installed ONCE, before rendering, with install_packages.R
# (see README -> "Quick start"). Rendering never installs anything itself.
needed_pkgs  <- c("ggplot2", "dplyr", "clusterProfiler", "biomaRt", "ggVennDiagram")
missing_pkgs <- needed_pkgs[!vapply(needed_pkgs, requireNamespace, logical(1),
                                    quietly = TRUE)]
if (length(missing_pkgs) > 0) {
  stop("Missing R packages: ", paste(missing_pkgs, collapse = ", "),
       "\nRun `Rscript install_packages.R` from the repository root first ",
       "(README -> Troubleshooting if Bioconductor installs fail).")
}

library(ggplot2)
library(dplyr)
library(clusterProfiler)
library(biomaRt)
library(ggVennDiagram)

cat("✅ All packages loaded\n")
```

---

# Load Data from Stage 2

```{r load-data}
# ── What we are loading ───────────────────────────────────────────────────
# DEG_results.RData    → degs_A, degs_B, results_A, results_B
# normalised_data.RData → norm_filtered (the full expression matrix with
#                         probe annotations in norm_filtered$genes)
# ─────────────────────────────────────────────────────────────────────────

results_path <- "results/"    # relative to the repository root (see README)
raw_path     <- "data/raw/"   # same folder as data_path in Stage 1

needed_files <- c(paste0(results_path, "DEG_results.RData"),
                  paste0(results_path, "normalised_data.RData"))
missing_files <- needed_files[!file.exists(needed_files)]
if (length(missing_files) > 0) {
  stop("Missing input file(s): ", paste(missing_files, collapse = ", "),
       "\nRender 01_load_QC_FIXED.Rmd and 02_differential_expression_FINAL.Rmd first (README -> How to run).")
}
if (length(list.files(raw_path, pattern = "\\.txt$")) == 0) {
  stop("No raw Agilent .txt files in '", raw_path, "'. Stage 3 re-reads one ",
       "of them for probe annotation (README -> Get the data).")
}

load(paste0(results_path, "DEG_results.RData"))
load(paste0(results_path, "normalised_data.RData"))

cat("✅ Data loaded successfully\n\n")
cat("Analysis A DEGs (Low P vs Reg P): ", nrow(degs_A), "\n")
cat("Analysis B DEGs (Low P vs Reg P, no outlier):", nrow(degs_B), "\n")
```

---

# Step 1: Extract PGSC Gene IDs

```{r extract-pgsc}
# ── What is a PGSC ID? ────────────────────────────────────────────────────
# PGSC stands for Potato Genome Sequencing Consortium.
# These IDs (e.g. PGSC0003DMT400069516) are the official potato gene
# identifiers from the 2011 genome assembly.
# DMT = transcript-level identifier
# DMG = gene-level identifier (same gene, different level)
# We extract these from the array annotation so we can look them up
# in external databases.
# ─────────────────────────────────────────────────────────────────────────

# ── Linking DEG rows to probe annotation ─────────────────────────────────
# rownames(degs_A) are the row names Stage 1 gave norm_filtered (assigned
# AFTER control probes and low-signal probes were removed). We MATCH them
# against rownames(norm_filtered$genes) rather than using them as numeric
# positions, and stop if anything fails to match — so a change in Stage 1
# filtering can never silently mis-align DEGs and annotation.
annot_for_degs <- function(deg_table) {
  idx <- match(rownames(deg_table), rownames(norm_filtered$genes))
  if (anyNA(idx)) {
    stop(sum(is.na(idx)), " DEG row names not found in norm_filtered$genes. ",
         "Re-render Stages 1 and 2 so both use the same filtered matrix.")
  }
  norm_filtered$genes[idx, ]
}

# Analysis A
annot_A       <- annot_for_degs(degs_A)
stopifnot(nrow(annot_A) == nrow(degs_A))

pgsc_ids_A <- unique(na.omit(annot_A$SystematicName))
pgsc_ids_A <- pgsc_ids_A[grepl("^PGSC", pgsc_ids_A)]  # keep PGSC IDs only (controls were already removed in Stage 1)

cat("✅ Analysis A: Extracted", length(pgsc_ids_A), "unique PGSC transcript IDs\n")
cat("Example IDs:\n")
print(head(pgsc_ids_A, 5))

# Analysis B
annot_B       <- annot_for_degs(degs_B)
pgsc_ids_B    <- unique(na.omit(annot_B$SystematicName))
pgsc_ids_B    <- pgsc_ids_B[grepl("^PGSC", pgsc_ids_B)]

cat("\n✅ Analysis B: Extracted", length(pgsc_ids_B), "unique PGSC transcript IDs\n")

# Save for reference
write.csv(data.frame(PGSC_ID = pgsc_ids_A),
          paste0(results_path, "DEG_PGSC_IDs_A.csv"), row.names = FALSE)
write.csv(data.frame(PGSC_ID = pgsc_ids_B),
          paste0(results_path, "DEG_PGSC_IDs_B.csv"), row.names = FALSE)

cat("\n✅ Saved PGSC ID lists to CSV\n")
```

---

# Step 2: Check Raw Agilent File Annotation

```{r check-raw-annotation}
# ── Why check the raw files? ──────────────────────────────────────────────
# The Agilent array files may contain extra annotation columns such as
# GeneName or Description that could help us identify genes.
# We check one file to see what columns are available.
# ─────────────────────────────────────────────────────────────────────────

raw_files <- list.files(raw_path, pattern = "\\.txt$", full.names = TRUE)
cat("Found", length(raw_files), "raw Agilent files\n")

# Function to read the data section of an Agilent file
read_agilent_features <- function(filepath) {
  lines       <- readLines(filepath, n = 15)
  header_line <- which(grepl("^FEATURES", lines))
  if (length(header_line) == 0) return(NULL)
  tryCatch(
    read.delim(filepath, skip = header_line - 1, header = TRUE,
               stringsAsFactors = FALSE, quote = ""),
    error = function(e) NULL
  )
}

first_raw <- read_agilent_features(raw_files[1])

if (!is.null(first_raw)) {
  cat("\nColumns in raw Agilent file:\n")
  print(colnames(first_raw))
  
  has_genename <- "GeneName"    %in% colnames(first_raw)
  has_desc     <- "Description" %in% colnames(first_raw)
  
  cat("\nGeneName column present:   ", has_genename, "\n")
  cat("Description column present:", has_desc, "\n")
}
```

```{r build-array-annotation}
# ── Build the full array annotation table ────────────────────────────────
# This contains one row per unique PGSC ID on the chip.
# It is kept for reference only. The enrichment background is NOT the
# whole chip: it is the set of probes actually tested (see Step 4).
# ─────────────────────────────────────────────────────────────────────────

if (!is.null(first_raw)) {
  
  anno_cols <- c("ProbeName", "SystematicName")
  if (has_genename) anno_cols <- c(anno_cols, "GeneName")
  if (has_desc)     anno_cols <- c(anno_cols, "Description")
  
  full_anno <- first_raw[first_raw$ControlType == 0, anno_cols]
  full_anno <- full_anno[grepl("^PGSC", full_anno$SystematicName), ]
  full_anno <- full_anno[!duplicated(full_anno$SystematicName), ]
  
  cat("Full array annotation: ", nrow(full_anno), "unique probe entries\n")
  cat("(Whole chip, for reference — the enrichment background is the tested set, Step 4)\n\n")
  
  write.csv(full_anno,
            paste0(results_path, "full_array_annotation.csv"), row.names = FALSE)
  cat("✅ Saved: full_array_annotation.csv\n")
  
} else {
  full_anno <- data.frame(ProbeName = character(), SystematicName = character())
  cat("⚠️ Could not build annotation table — check raw_path\n")
}
```

---

# Step 3: Retrieve GO Annotations via Ensembl Plants BioMart

```{r biomart-connect}
# ── What is BioMart? ──────────────────────────────────────────────────────
# BioMart is Ensembl's online database query tool.
# We connect to the Ensembl Plants version, which hosts potato genome data.
# From here we can retrieve GO term annotations for our PGSC transcript IDs.
#
# Note: This requires an internet connection. It may take 1–2 minutes.
# ─────────────────────────────────────────────────────────────────────────

cat("Connecting to Ensembl Plants BioMart...\n")
cat("(This requires internet — may take 1-2 minutes)\n\n")

plants_mart <- useMart(
  biomart = "plants_mart",
  dataset = "stuberosum_eg_gene",
  host    = "https://plants.ensembl.org"
)

cat("✅ Connected to Ensembl Plants\n")
cat("Dataset: Solanum tuberosum genes (DM v6.1)\n")
```

```{r download-potato-annotation}
# ── Download GO annotations for ALL potato genes ─────────────────────────
# We download the full annotation (not just our DEGs) because we need
# to know what GO terms exist across the whole genome — that is our
# statistical background.
#
# What we retrieve:
# ensembl_gene_id   = PGSC gene ID (DMG format)
# go_id             = GO term accession (e.g. GO:0006796)
# name_1006         = GO term name (e.g. "phosphate metabolic process")
# namespace_1003    = GO domain (biological_process, molecular_function,
#                     cellular_component)
# ─────────────────────────────────────────────────────────────────────────

cat("Downloading potato GO annotations from Ensembl Plants...\n")
cat("(This is a large download — may take 2-3 minutes)\n\n")

potato_anno <- getBM(
  attributes = c(
    "ensembl_gene_id",
    "external_gene_name",
    "go_id",
    "name_1006",
    "namespace_1003"
  ),
  mart = plants_mart
)

cat("✅ Download complete\n")
cat("Total annotation rows:", nrow(potato_anno), "\n")
cat("Unique potato genes:  ", length(unique(potato_anno$ensembl_gene_id)), "\n\n")

# Save so you don't have to re-download every session
write.csv(potato_anno,
          paste0(results_path, "potato_ensembl_annotation.csv"),
          row.names = FALSE)
cat("✅ Saved: potato_ensembl_annotation.csv\n")
cat("   (Load this file next time instead of re-downloading)\n")
```

```{r convert-ids}
# ── Why do we convert DMT to DMG? ────────────────────────────────────────
# Our array uses TRANSCRIPT-level IDs (DMT: ...DMT400XXXXXX)
# Ensembl Plants stores GENE-level IDs   (DMG: ...DMG400XXXXXX)
# They share the same number — only the letter changes.
# So PGSC0003DMT400069516 → PGSC0003DMG400069516
# This lets us match our DEGs to the Ensembl annotation.
# ─────────────────────────────────────────────────────────────────────────

# Convert both DEG lists and the full array
pgsc_ids_A_gene  <- gsub("DMT", "DMG", pgsc_ids_A)
pgsc_ids_B_gene  <- gsub("DMT", "DMG", pgsc_ids_B)
full_anno_gene   <- gsub("DMT", "DMG", full_anno$SystematicName)

cat("ID conversion: DMT → DMG\n")
cat("Example: PGSC0003DMT400069516 → PGSC0003DMG400069516\n\n")

# Filter annotation to our DEGs
degs_anno_A <- potato_anno[potato_anno$ensembl_gene_id %in% pgsc_ids_A_gene, ]
has_go_A    <- degs_anno_A[degs_anno_A$go_id != "", ]

cat("=== ANNOTATION COVERAGE (Analysis A) ===\n")
cat("Total DEGs:              ", length(pgsc_ids_A_gene), "\n")
cat("Matched to Ensembl:      ", length(unique(degs_anno_A$ensembl_gene_id)), "\n")
cat("With at least one GO term:", length(unique(has_go_A$ensembl_gene_id)), "\n\n")

# Computed, not typed in: fraction of DEG genes with at least one GO term
n_deg_genes_A <- length(unique(pgsc_ids_A_gene))
n_go_A        <- length(unique(has_go_A$ensembl_gene_id))
go_cov_A      <- n_go_A / n_deg_genes_A
cat("GO coverage of Analysis A DEGs:", round(100 * go_cov_A, 1), "%",
    "(", n_go_A, "of", n_deg_genes_A, ")\n")
cat("Low coverage is a known limitation of PGSC v3.4 / non-model annotation.\n")
```

---

# Step 4: Build Background Gene Set

```{r build-background}
# ── What is a background gene set and why does it matter? ────────────────
# The background (also called the "universe") is the set of ALL genes
# that COULD have been in our DEG list — i.e. every gene that was actually
# TESTED in Stage 2: probes that passed the Stage 1 expression filter, with
# Agilent control probes removed. Genes on the chip that were filtered out
# (not expressed in roots) could never have become DEGs, so including them
# would inflate enrichment of root-expressed functions.
#
# If we don't specify this, enricher() defaults to every gene in TERM2GENE.
# (An even worse mistake is using the DEG list as its own background, which
# makes GeneRatio == BgRatio and all p-values = 1.)
#
# The tested IDs are mapped exactly like the DEG IDs:
# norm_filtered$genes$SystematicName → PGSC IDs only → DMT → DMG.
# ─────────────────────────────────────────────────────────────────────────

tested_pgsc <- unique(na.omit(norm_filtered$genes$SystematicName))
tested_pgsc <- tested_pgsc[grepl("^PGSC", tested_pgsc)]
tested_gene <- unique(gsub("DMT", "DMG", tested_pgsc))

# Every DEG must be in the tested set (it was tested by definition)
stopifnot(all(pgsc_ids_A_gene %in% tested_gene),
          all(pgsc_ids_B_gene %in% tested_gene))

# Whole-array GO annotation, kept for reference / comparison only
full_array_anno <- potato_anno[
  potato_anno$ensembl_gene_id %in% full_anno_gene, ]

# GO annotation of the TESTED genes — this is the enrichment background
tested_anno <- potato_anno[potato_anno$ensembl_gene_id %in% tested_gene, ]

# Build term2gene: maps each GO term to the genes that carry it
# This is the format clusterProfiler's enricher() needs.
# (Object name kept from earlier versions; it is now restricted to the
#  tested genes, not the full array.)
term2gene_full <- tested_anno[tested_anno$go_id != "",
                              c("go_id", "ensembl_gene_id")]
colnames(term2gene_full) <- c("GO", "gene")

# Build term2name: maps each GO ID to its human-readable name
term2name_full <- unique(tested_anno[tested_anno$go_id != "",
                                     c("go_id", "name_1006")])
colnames(term2name_full) <- c("GO", "name")

# Effective background size = tested genes that carry >= 1 GO term
n_bg_go <- length(unique(term2gene_full$gene))

cat("=== BACKGROUND GENE SET (tested genes) ===\n")
cat("Whole-array genes with Ensembl annotation (reference):",
    length(unique(full_array_anno$ensembl_gene_id)), "\n")
cat("Tested genes (passed Stage 1 filter, controls removed):", length(tested_gene), "\n")
cat("Tested genes with Ensembl annotation:                  ",
    length(unique(tested_anno$ensembl_gene_id)), "\n")
cat("Tested genes with >= 1 GO term (effective background): ", n_bg_go, "\n")
cat("GO term–gene pairs in background:                      ", nrow(term2gene_full), "\n")
cat("Unique GO terms in background:                         ",
    length(unique(term2gene_full$GO)), "\n\n")
cat("✅ Background set to the tested genes (passed as `universe`)\n")
```

---

# Step 5: GO Enrichment — Analysis A

```{r go-enrichment-A}
# ── Running the enrichment test ───────────────────────────────────────────
# enricher() from clusterProfiler runs a hypergeometric test.
# It asks: given that X genes in the background carry GO term Y,
# what is the probability of seeing at least Z of our DEGs with that term?
#
# Parameters explained:
# gene          = our DEG list (query)
# universe      = the background: genes actually tested (Step 4)
# TERM2GENE     = GO term → gene mapping (restricted to tested genes)
# TERM2NAME     = GO term ID → readable name
# pvalueCutoff  = only show terms with p < 0.05
# pAdjustMethod = "BH" = Benjamini-Hochberg correction for multiple testing
# minGSSize     = minimum number of genes a GO term must have to be tested
# ─────────────────────────────────────────────────────────────────────────

cat("Running GO enrichment for Analysis A...\n\n")

ego_A <- enricher(
  gene          = pgsc_ids_A_gene,
  universe      = tested_gene,     # background = genes actually tested
  TERM2GENE     = term2gene_full,
  TERM2NAME     = term2name_full,
  pvalueCutoff  = 0.05,
  pAdjustMethod = "BH",
  minGSSize     = 3,
  maxGSSize     = 500
)

# ── Interpret results ─────────────────────────────────────────────────────
sig_A <- ego_A@result[ego_A@result$p.adjust < 0.05, ]
nom_A <- ego_A@result[ego_A@result$pvalue   < 0.05 &
                      nchar(trimws(ego_A@result$Description)) > 0, ]
nom_A <- nom_A[order(nom_A$pvalue), ]

cat("GO terms tested:              ", nrow(ego_A@result), "\n")
cat("Significant (BH adj.p < 0.05):", nrow(sig_A), "\n")
cat("Nominal p < 0.05:             ", nrow(nom_A), "\n\n")

# ── Why no BH-corrected results? (message computed from the results) ─────
if (nrow(sig_A) == 0) {
  cat("Note: No terms survive BH correction.\n")
  cat("This is a STATISTICAL POWER issue, not a code error.\n")
} else {
  cat("Note:", nrow(sig_A), "term(s) survive BH correction (see sig_A).\n")
}
cat("Only", n_go_A, "of", n_deg_genes_A, "Analysis A DEG genes carry GO annotations.\n")
cat("The hypergeometric test has limited power with a small annotated set.\n")
cat("Reporting nominal p < 0.05 as exploratory findings is standard\n")
cat("practice in non-model organism transcriptomics.\n\n")

cat("Top nominal results:\n")
print(head(nom_A[, c("Description", "GeneRatio", "BgRatio",
                      "pvalue", "p.adjust", "Count")], 15))
```

---

# Step 6: GO Enrichment — Analysis B

```{r go-enrichment-B}
# Same approach for Analysis B (Low P vs Reg P, outlier removed)

cat("Running GO enrichment for Analysis B...\n\n")

degs_anno_B <- potato_anno[potato_anno$ensembl_gene_id %in% pgsc_ids_B_gene, ]
has_go_B    <- degs_anno_B[degs_anno_B$go_id != "", ]

cat("=== ANNOTATION COVERAGE (Analysis B) ===\n")
cat("Total DEGs:               ", length(pgsc_ids_B_gene), "\n")
cat("Matched to Ensembl:       ",
    length(unique(degs_anno_B$ensembl_gene_id)), "\n")
cat("With at least one GO term:",
    length(unique(has_go_B$ensembl_gene_id)), "\n\n")

ego_B <- enricher(
  gene          = pgsc_ids_B_gene,
  universe      = tested_gene,     # same background as Analysis A
  TERM2GENE     = term2gene_full,
  TERM2NAME     = term2name_full,
  pvalueCutoff  = 0.05,
  pAdjustMethod = "BH",
  minGSSize     = 3,
  maxGSSize     = 500
)

nom_B <- ego_B@result[ego_B@result$pvalue < 0.05 &
                      nchar(trimws(ego_B@result$Description)) > 0, ]
nom_B <- nom_B[order(nom_B$pvalue), ]

cat("GO terms tested (B):              ", nrow(ego_B@result), "\n")
cat("BH-significant (B):               ",
    sum(ego_B@result$p.adjust < 0.05), "\n")
cat("Nominal p < 0.05 (B):             ", nrow(nom_B), "\n\n")

cat("Top nominal results (B):\n")
print(head(nom_B[, c("Description", "GeneRatio", "BgRatio",
                      "pvalue", "Count")], 10))
```

---

# Step 7: Visualise GO Enrichment

## Parse gene ratio for plotting

```{r parse-ratio}
# ── What is GeneRatio? ────────────────────────────────────────────────────
# GeneRatio is stored as a string like "3/76", which would mean
# "3 of our 76 annotated DEGs carry this GO term".
# We convert it to a decimal so we can use it for dot size/position.
# ─────────────────────────────────────────────────────────────────────────

parse_ratio <- function(x) {
  parts <- strsplit(x, "/")[[1]]
  as.numeric(parts[1]) / as.numeric(parts[2])
}

# Prepare Analysis A plot data
nom_A_plot <- nom_A[nchar(trimws(nom_A$Description)) > 0, ]
nom_A_plot$GeneRatioNum <- sapply(nom_A_plot$GeneRatio, parse_ratio)
nom_A_plot$ShortDesc <- ifelse(
  nchar(nom_A_plot$Description) > 50,
  paste0(substr(nom_A_plot$Description, 1, 47), "..."),
  nom_A_plot$Description
)

# Prepare Analysis B plot data
nom_B_plot <- nom_B[nchar(trimws(nom_B$Description)) > 0, ]
nom_B_plot$GeneRatioNum <- sapply(nom_B_plot$GeneRatio, parse_ratio)
nom_B_plot$ShortDesc <- ifelse(
  nchar(nom_B_plot$Description) > 50,
  paste0(substr(nom_B_plot$Description, 1, 47), "..."),
  nom_B_plot$Description
)

cat("Plot data prepared\n")
cat("Analysis A terms to plot:", nrow(nom_A_plot), "\n")
cat("Analysis B terms to plot:", nrow(nom_B_plot), "\n")
```

## GO dotplot — Analysis A

```{r go-dotplot-A}
# ── How to read this dotplot ──────────────────────────────────────────────
# Each row = one GO term
# X-axis = Gene Ratio (what fraction of our DEGs carry this term)
# Dot size = number of DEG genes in that term
# Dot colour = nominal p-value (darker = more significant)
# ─────────────────────────────────────────────────────────────────────────

p_go_A <- ggplot(nom_A_plot,
                 aes(x      = GeneRatioNum,
                     y      = reorder(ShortDesc, -pvalue),
                     size   = Count,
                     colour = pvalue)) +
  geom_point(alpha = 0.85) +
  scale_colour_gradient(low  = "#1a6b8a", high = "#d4e8f0",
                        name = "Nominal p-value") +
  scale_size_continuous(name = "Gene Count", range = c(3, 10)) +
  theme_minimal(base_size = 12) +
  labs(
    title    = "GO Enrichment — Analysis A",
    subtitle = paste0(
      "Low Phosphate vs Regular Phosphate | Nominal p < 0.05\n",
      "(", ifelse(nrow(sig_A) == 0, "BH correction not significant",
                  paste(nrow(sig_A), "term(s) BH-significant")),
      " — ", n_go_A, "/", n_deg_genes_A, " DEG genes annotated, ",
      "PGSC v3.4 sparse annotation)"
    ),
    x = "Gene Ratio",
    y = NULL
  ) +
  theme(
    plot.title    = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(size  = 9, colour = "grey40"),
    axis.text.y   = element_text(size  = 10)
  )

print(p_go_A)

ggsave(paste0(results_path, "GO_dotplot_Analysis_A.png"),
       p_go_A, width = 10, height = 8, dpi = 300)
cat("✅ Saved: GO_dotplot_Analysis_A.png\n")
```

## GO dotplot — Analysis B

```{r go-dotplot-B}
p_go_B <- ggplot(nom_B_plot,
                 aes(x      = GeneRatioNum,
                     y      = reorder(ShortDesc, -pvalue),
                     size   = Count,
                     colour = pvalue)) +
  geom_point(alpha = 0.85) +
  scale_colour_gradient(low  = "#8a1a1a", high = "#f0d4d4",
                        name = "Nominal p-value") +
  scale_size_continuous(name = "Gene Count", range = c(3, 10)) +
  theme_minimal(base_size = 12) +
  labs(
    title    = "GO Enrichment — Analysis B",
    subtitle = "Low Phosphate vs Regular Phosphate (outlier removed) | Nominal p < 0.05",
    x        = "Gene Ratio",
    y        = NULL
  ) +
  theme(
    plot.title    = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(size  = 9, colour = "grey40"),
    axis.text.y   = element_text(size  = 10)
  )

print(p_go_B)

ggsave(paste0(results_path, "GO_dotplot_Analysis_B.png"),
       p_go_B, width = 10, height = 8, dpi = 300)
cat("✅ Saved: GO_dotplot_Analysis_B.png\n")
```

---

# Step 8: DEG Overlap — Analysis A vs B

```{r deg-overlap}
# ── What does the overlap tell us? ───────────────────────────────────────
# Analysis A and B test the SAME contrast (Low P vs Reg P); B simply drops
# the outlier Reg P replicate. A gene that is a DEG in BOTH analyses is
# therefore robust to that replicate — a stronger, more reliable finding.
# ─────────────────────────────────────────────────────────────────────────

overlap_genes <- intersect(pgsc_ids_A_gene, pgsc_ids_B_gene)
only_A        <- setdiff(pgsc_ids_A_gene, pgsc_ids_B_gene)
only_B        <- setdiff(pgsc_ids_B_gene, pgsc_ids_A_gene)

cat("=== DEG OVERLAP ===\n")
cat("Analysis A only:", length(only_A), "genes\n")
cat("Analysis B only:", length(only_B), "genes\n")
cat("Shared (A ∩ B): ", length(overlap_genes), "genes\n")
cat("Overlap %:      ",
    round(100 * length(overlap_genes) / length(pgsc_ids_B_gene), 1),
    "% of Analysis B DEGs also in Analysis A\n\n")
cat("Interpretation: a high overlap indicates that the core transcriptional\n")
cat("response does not depend on the outlier Reg P replicate.\n")
```

```{r venn-ab}
venn_list <- list(
  "Analysis A\n(Low P vs Reg P)"  = pgsc_ids_A_gene,
  "Analysis B\n(Low P vs Reg P,\nno outlier)" = pgsc_ids_B_gene
)

p_venn <- ggVennDiagram(
  venn_list,
  label_alpha = 0,
  label       = "count",
  set_label_size = 3.5
) +
  scale_fill_gradient(low = "#d4e8f0", high = "#1a6b8a") +
  scale_colour_manual(values = c("grey30", "grey30")) +
  labs(
    title    = "DEG Overlap — Analysis A vs Analysis B",
    subtitle = "Maris Piper | E-MTAB-2990"
  ) +
  theme(
    plot.title      = element_text(face = "bold", size = 13, hjust = 0.5),
    plot.subtitle   = element_text(size = 10, hjust = 0.5, colour = "grey40"),
    legend.position = "none",
    plot.margin     = margin(10, 30, 10, 30)
  )

print(p_venn)

ggsave(paste0(results_path, "Venn_DEG_overlap_AB.png"),
       p_venn, width = 8, height = 7, dpi = 300)
cat("✅ Saved: Venn_DEG_overlap_AB.png\n")
```

## GO term overlap — A vs B

```{r go-overlap}
# ── Why look at shared GO terms? ──────────────────────────────────────────
# If the same GO term appears as nominally enriched in BOTH analyses
# independently, that is much stronger evidence than a single analysis.
# It means the biological signal is reproducible.
# ─────────────────────────────────────────────────────────────────────────

go_A_terms <- nom_A_plot$Description
go_B_terms <- nom_B_plot$Description
shared_go  <- intersect(go_A_terms, go_B_terms)

cat("=== GO TERM OVERLAP ===\n")
cat("Nominal GO terms in A:", length(go_A_terms), "\n")
cat("Nominal GO terms in B:", length(go_B_terms), "\n")
cat("Shared GO terms:      ", length(shared_go),  "\n\n")

cat("Shared terms:\n")
for (i in seq_along(shared_go)) {
  cat(sprintf("  %2d. %s\n", i, shared_go[i]))
}
```

---

# Step 9: Key Gene Annotation Lookup

```{r key-gene-annotation}
# ── What are we doing here? ───────────────────────────────────────────────
# GO enrichment tells us WHICH biological processes are enriched, but
# it doesn't tell us the individual gene names.
# Here we query BioMart directly to get the description of each DEG,
# then search those descriptions for biologically relevant keywords.
# This gives us named, annotated gene lists for each functional category.
# ─────────────────────────────────────────────────────────────────────────

cat("Fetching full gene descriptions for all", length(pgsc_ids_A),
    "Analysis A DEG transcripts...\n")

full_deg_anno <- getBM(
  attributes = c(
    "ensembl_transcript_id",
    "ensembl_gene_id",
    "external_gene_name",
    "description",
    "go_id",
    "name_1006",
    "namespace_1003"
  ),
  filters = "ensembl_transcript_id",
  values  = pgsc_ids_A,
  mart    = plants_mart
)

cat("✅ Retrieved", nrow(full_deg_anno), "annotation rows\n")
cat("Unique genes with any annotation:",
    length(unique(full_deg_anno$ensembl_transcript_id)), "\n\n")
```

```{r key-gene-categories}
# ── Keyword search across functional categories ───────────────────────────
# We define biologically relevant categories for phosphate stress
# and search gene descriptions for matching keywords.
# ─────────────────────────────────────────────────────────────────────────

key_terms <- list(
  acid_phosphatase    = c("acid phosphatase", "purple acid phosphatase", "PAP"),
  phosphate_transport = c("phosphate transporter", "PHT", "inorganic phosphate"),
  cell_wall           = c("cell wall", "extensin", "expansin", "pectin",
                          "xyloglucan", "polygalacturonase", "peroxidase"),
  auxin               = c("auxin", "IAA", "SAUR", "indole"),
  stress_response     = c("stress", "dehydration", "defense", "IDS4",
                          "MLO", "SPX", "ethylene-responsive"),
  transcription       = c("transcription factor", "WRKY", "MYB", "NAC",
                          "bHLH", "ERF", "AP2", "zinc finger", "GRAS"),
  lipid               = c("lipid", "fatty acid", "phospholipid",
                          "galactolipid", "glycerophosphodiester",
                          "acyl", "diacylglycerol"),
  membrane_transport  = c("transporter", "aquaporin", "channel",
                          "ABC transporter", "nodulin")
)

results_list <- list()

for (term_name in names(key_terms)) {
  keywords <- key_terms[[term_name]]
  pattern  <- paste(keywords, collapse = "|")
  hits <- full_deg_anno[
    grepl(pattern, full_deg_anno$description, ignore.case = TRUE) |
    grepl(pattern, full_deg_anno$name_1006,   ignore.case = TRUE), ]
  if (nrow(hits) > 0) {
    hits_unique            <- hits[!duplicated(hits$ensembl_transcript_id), ]
    hits_unique$Category   <- term_name
    results_list[[term_name]] <- hits_unique
    cat("✅", sprintf("%-22s", term_name), ":", nrow(hits_unique), "genes\n")
  } else {
    cat("—", sprintf("%-22s", term_name), ": 0 genes\n")
  }
}
```

```{r expression-direction}
# ── Add expression direction ──────────────────────────────────────────────
# For each annotated gene, we look up its logFC from Stage 2 results.
# This tells us whether it was UPREGULATED or DOWNREGULATED under
# low phosphate conditions.
# ─────────────────────────────────────────────────────────────────────────

cat("\n=== EXPRESSION DIRECTION FOR KEY GENE FAMILIES ===\n\n")

for (term_name in names(results_list)) {
  cat("---", term_name, "---\n")
  hits <- results_list[[term_name]]
  for (i in seq_len(nrow(hits))) {
    tid       <- hits$ensembl_transcript_id[i]
    # Match against annot_A (row-aligned with degs_A), NOT pgsc_ids_A:
    # pgsc_ids_A is de-duplicated and has control probes removed, so its
    # positions do not correspond to rows of degs_A.
    match_idx <- which(annot_A$SystematicName == tid)
    if (length(match_idx) > 0) {
      lfc <- degs_A$logFC[match_idx[1]]
      dir <- ifelse(lfc > 0, "⬆ UP", "⬇ DOWN")
      desc_clean <- gsub("\\s*\\[Source:.*\\]", "", hits$description[i])
      cat(sprintf("  %s | logFC = %+.2f %s | %s\n",
                  tid, lfc, dir,
                  substr(desc_clean, 1, 55)))
    }
  }
  cat("\n")
}

# Combine and save
all_hits <- do.call(rbind, results_list)
write.csv(all_hits,
          paste0(results_path, "Key_genes_annotated.csv"), row.names = FALSE)
cat("✅ Saved: Key_genes_annotated.csv\n")
```

---

# Step 10: Functional Category Summary Plot

```{r category-summary-plot}
# ── Interpreting the category counts ─────────────────────────────────────
# These numbers come directly from the BioMart keyword search above.
# The direction label summarises whether the genes in each category
# are all/mostly upregulated, all/mostly downregulated, or mixed.
# ─────────────────────────────────────────────────────────────────────────

# Counts and directions are COMPUTED from results_list (keyword hits) and
# the Analysis A logFC of each hit — nothing is typed in.
# Count     = number of unique DEG transcripts matched to the category
# Direction = from the signs of their logFC (same probe match as the
#             expression-direction chunk: first matching DEG probe):
#   100% up -> "ALL UP"       >= 2/3 up -> "Mostly UP"
#   100% down -> "ALL DOWN"   <= 1/3 up -> "Mostly DOWN"   otherwise "Mixed"
cat_display <- c(
  membrane_transport  = "Membrane Transport",
  lipid               = "Lipid Metabolism",
  stress_response     = "Stress Response",
  cell_wall           = "Cell Wall",
  transcription       = "Transcription Factors",
  acid_phosphatase    = "Acid Phosphatase",
  phosphate_transport = "Phosphate Transport",
  auxin               = "Auxin Signalling"
)

direction_label <- function(lfc) {
  lfc <- lfc[!is.na(lfc)]
  if (length(lfc) == 0) return(NA_character_)
  frac_up <- mean(lfc > 0)
  if (frac_up == 1)        "ALL UP"
  else if (frac_up == 0)   "ALL DOWN"
  else if (frac_up >= 2/3) "Mostly UP"
  else if (frac_up <= 1/3) "Mostly DOWN"
  else                     "Mixed"
}

cat_summary <- do.call(rbind, lapply(names(results_list), function(k) {
  hits <- results_list[[k]]
  lfc  <- degs_A$logFC[match(hits$ensembl_transcript_id, annot_A$SystematicName)]
  data.frame(
    Key       = k,
    Category  = unname(cat_display[k]),
    Count     = nrow(hits),
    N_up      = sum(lfc > 0, na.rm = TRUE),
    N_down    = sum(lfc < 0, na.rm = TRUE),
    Direction = direction_label(lfc),
    stringsAsFactors = FALSE
  )
}))
if (is.null(cat_summary)) {
  stop("No DEG matched any functional keyword category — nothing to plot.")
}
cat_summary <- cat_summary[order(-cat_summary$Count), ]
print(cat_summary, row.names = FALSE)

p_cats <- ggplot(cat_summary,
                 aes(x    = reorder(Category, Count),
                     y    = Count,
                     fill = Direction)) +
  geom_col(alpha = 0.85, width = 0.7) +
  scale_fill_manual(values = c(
    "ALL UP"      = "#C0392B",
    "Mostly UP"   = "#E67E22",
    "Mixed"       = "#7F8C8D",
    "Mostly DOWN" = "#5DADE2",
    "ALL DOWN"    = "#2471A3"
  )) +
  coord_flip() +
  theme_minimal(base_size = 12) +
  labs(
    title    = "DEGs by Functional Category — Analysis A",
    subtitle = "Maris Piper | Low Phosphate vs Regular Phosphate\nColour = expression direction under low phosphate",
    x        = NULL,
    y        = "Number of DEGs",
    fill     = "Expression Direction"
  ) +
  theme(
    plot.title      = element_text(face = "bold"),
    legend.position = "bottom"
  )

print(p_cats)

ggsave(paste0(results_path, "Functional_categories_barplot.png"),
       p_cats, width = 9, height = 6, dpi = 300)
cat("✅ Saved: Functional_categories_barplot.png\n")
```

---

# Step 11: Results Summary

```{r results-summary}
cat("╔══════════════════════════════════════════════════════╗\n")
cat("║           STAGE 3 — RESULTS SUMMARY                 ║\n")
cat("╠══════════════════════════════════════════════════════╣\n")
cat(sprintf("║  Analysis A DEGs (Low P vs Reg P):     %4d          ║\n",
            length(pgsc_ids_A)))
cat(sprintf("║  Analysis B DEGs (no outlier):         %4d          ║\n",
            length(pgsc_ids_B)))
cat(sprintf("║  Shared DEGs (A ∩ B):                  %4d  (%2.0f%%)   ║\n",
            length(overlap_genes),
            100 * length(overlap_genes) / length(pgsc_ids_B_gene)))
cat("╠══════════════════════════════════════════════════════╣\n")
cat("║  KEGG (direct):         ❌ PGSC IDs not in KEGG      ║\n")
cat("║  KEGG (Arabidopsis):    ❌ 151 orthologs, no sig.     ║\n")
cat("║    (author's notes, outside notebook — to verify)    ║\n")
cat(sprintf("║  GO Analysis A (nominal): ✅ %2d terms (p<0.05)        ║\n",
            nrow(nom_A_plot)))
cat(sprintf("║  GO Analysis B (nominal): ✅ %2d terms (p<0.05)        ║\n",
            nrow(nom_B_plot)))
cat(sprintf("║  Shared GO terms:         ✅ %2d terms both analyses   ║\n",
            length(shared_go)))
cat("╠══════════════════════════════════════════════════════╣\n")
cat("║  KEY BIOLOGICAL FINDINGS (computed):                 ║\n")

# Category findings, straight from cat_summary
cat_line <- function(key, label) {
  row <- cat_summary[cat_summary$Key == key, ]
  if (nrow(row) == 0) return(sprintf("0 %s", label))
  sprintf("%d %s — %s (%d up / %d down)",
          row$Count, label, row$Direction, row$N_up, row$N_down)
}
findings <- c(
  cat_line("acid_phosphatase",    "acid phosphatases"),
  cat_line("phosphate_transport", "phosphate transporters"),
  cat_line("auxin",               "auxin-related"),
  cat_line("lipid",               "lipid-related")
)

# Root hair GO terms: nominal in A, B, both?
rh_A <- any(grepl("root hair", go_A_terms, ignore.case = TRUE))
rh_B <- any(grepl("root hair", go_B_terms, ignore.case = TRUE))
findings <- c(findings, paste0("Root hair GO term nominal in: ",
  if (rh_A && rh_B) "A and B" else if (rh_A) "A only" else if (rh_B) "B only" else "neither"))

# Top DEG = largest |logFC| in Analysis A, with its BioMart description
top_i    <- which.max(abs(degs_A$logFC))
top_id   <- annot_A$SystematicName[top_i]
top_desc <- full_deg_anno$description[match(top_id, full_deg_anno$ensembl_transcript_id)]
top_desc <- if (is.na(top_desc) || !nzchar(top_desc)) "no BioMart description" else
            substr(gsub("\\s*\\[Source:.*\\]", "", top_desc), 1, 40)
findings <- c(findings,
  sprintf("Top DEG (max |logFC|): %s, logFC = %+.2f", top_id, degs_A$logFC[top_i]),
  sprintf("  -> %s", top_desc))

for (f in findings) cat(sprintf("║  • %s\n", f))
cat("╚══════════════════════════════════════════════════════╝\n")
```

---

# Step 12: Save All Results

```{r save-all}
cat("=== SAVING ALL STAGE 3 OUTPUTS ===\n\n")

# GO enrichment results
write.csv(nom_A[, c("ID", "Description", "GeneRatio", "BgRatio",
                    "pvalue", "p.adjust", "Count")],
          paste0(results_path, "GO_enrichment_nominal_A.csv"),
          row.names = FALSE)

write.csv(nom_B[, c("ID", "Description", "GeneRatio", "BgRatio",
                    "pvalue", "p.adjust", "Count")],
          paste0(results_path, "GO_enrichment_nominal_B.csv"),
          row.names = FALSE)

# Save R objects
objects_to_save <- c(
  "pgsc_ids_A", "pgsc_ids_B",
  "pgsc_ids_A_gene", "pgsc_ids_B_gene",
  "ego_A", "ego_B",
  "nom_A", "nom_B",
  "overlap_genes", "shared_go",
  "full_deg_anno", "full_anno",
  "term2gene_full", "term2name_full",
  "tested_gene",
  "cat_summary"
)

objects_to_save <- objects_to_save[sapply(objects_to_save, exists)]

save(list = objects_to_save,
     file = paste0(results_path, "Stage3_complete.RData"))

cat("✅ Files saved:\n")
cat("  GO_enrichment_nominal_A.csv\n")
cat("  GO_enrichment_nominal_B.csv\n")
cat("  Key_genes_annotated.csv\n")
cat("  potato_ensembl_annotation.csv\n")
cat("  GO_dotplot_Analysis_A.png\n")
cat("  GO_dotplot_Analysis_B.png\n")
cat("  Venn_DEG_overlap_AB.png\n")
cat("  Functional_categories_barplot.png\n")
cat("  Stage3_complete.RData\n\n")
cat("🎯 Stage 3 complete! Proceed to Stage 4.\n")
```

---

# Methods Section (Copy Into Dissertation)

## Methods text

```{r methods-numbers, include=FALSE}
# Numbers used in the Methods paragraph below — all computed from the objects
# above, so the text stays in step with the data.
m_n_deg_probes <- nrow(degs_A)
m_n_deg_tx     <- length(pgsc_ids_A)
m_n_tested     <- length(tested_gene)
m_n_bg_go      <- n_bg_go
m_go_cov_pct   <- round(100 * go_cov_A)
m_bh_text      <- if (nrow(sig_A) == 0) {
  "but no terms survived at the adjusted p < 0.05 threshold. This is attributable to the sparse GO annotation coverage of the PGSC v3.4 assembly"
} else {
  paste0("and ", nrow(sig_A), " term(s) survived at the adjusted p < 0.05 threshold. GO annotation coverage of the PGSC v3.4 assembly was nevertheless sparse")
}
m_v_biomart    <- as.character(packageVersion("biomaRt"))
m_v_cp         <- as.character(packageVersion("clusterProfiler"))
```

> Functional enrichment analysis was performed on `r m_n_deg_probes` DEG probes (`r m_n_deg_tx` unique PGSC transcripts) identified in Analysis A (Low Phosphate vs Regular Phosphate). Agilent control probes had been removed in Stage 1, before differential expression testing. PGSC transcript identifiers (format: DMT) were converted to gene-level identifiers (DMG) for compatibility with Ensembl Plants BioMart. Gene Ontology (GO) annotations were retrieved from Ensembl Plants BioMart (Solanum tuberosum, DM v6.1 assembly) using the R package biomaRt (v`r m_v_biomart`). GO enrichment was performed using the enricher() function in clusterProfiler (v`r m_v_cp`; Wu et al., 2021), with the genes actually tested as the statistical background (`r m_n_tested` genes that passed Stage 1 expression filtering, of which `r m_n_bg_go` carried at least one GO term). Significance was assessed by hypergeometric test; Benjamini-Hochberg multiple testing correction was applied `r m_bh_text` (approximately `r m_go_cov_pct`% of DEG genes carried GO annotations). Results are therefore reported at nominal p < 0.05 as exploratory findings, consistent with standard practice in non-model organism transcriptomics (Frades et al., 2015). KEGG pathway enrichment was attempted outside this notebook via direct PGSC-to-KEGG ID mapping and via Arabidopsis thaliana ortholog mapping (151 high-confidence orthologs, >40% identity — *to verify after re-run*; this step is not reproducible from the repository); neither approach yielded significant results. Individual gene annotation was performed by querying gene descriptions from BioMart and categorising DEGs by keyword matching.

## References to cite

- Wu T et al. (2021) clusterProfiler 4.0. *Innovation* 2(3):100141
- PGSC (2011) Genome sequence and analysis of the tuber crop potato. *Nature* 475:189–195
- Frades I et al. (2015) Novel hypothesis on how freezing affects potato. *Front Plant Sci* 6:583



# Session Info

```{r session}
sessionInfo()
```
