---
title: "Stage 4: Advanced Analysis & Cross-Cultivar Comparison"
subtitle: "Maris Piper Potato — Phosphate Stress Transcriptomics | E-MTAB-2990"
author: "MSc Bioinformatics Internship"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 7)
```

---

# Overview & Learning Goals

This notebook is the final analytical stage of the E-MTAB-2990 phosphate stress project.
By the end of this stage you will have:

- Compared how all 4 potato cultivars respond to low phosphate (cross-cultivar analysis)
- Identified which Maris Piper genes are unique vs shared across cultivars
- Produced 10 publication-quality figures for your report/presentation
- Understood the biological story told by the data

**What is this dataset?**
24 Agilent microarray samples from 4 potato cultivars (Maris Piper, Stirling, Pentland Dell, 12601),
each grown under 2 phosphate conditions: Low P (phosphate starvation) and Regular P (normal).
3 biological replicates per group. We already identified 962 DEGs in Maris Piper (Stage 2).
Now we ask: are these genes specific to Maris Piper, or do all cultivars respond the same way?

---

# Load Libraries & Data

```{r load}
# --- Core libraries ---
library(limma)       # Microarray statistics (DEG analysis)
library(ggplot2)     # Plotting
library(dplyr)       # Data manipulation
library(pheatmap)    # Heatmaps
library(RColorBrewer) # Colour palettes

# --- Extra libraries (install if needed) ---
if (!requireNamespace("ggrepel",       quietly = TRUE)) install.packages("ggrepel")
if (!requireNamespace("ggVennDiagram", quietly = TRUE)) install.packages("ggVennDiagram")
library(ggrepel)       # Smart non-overlapping labels on plots
library(ggVennDiagram) # 4-way Venn diagrams

# --- Set your results folder path ---
# This is where Stage 1-3 outputs were saved, and where Stage 4 will also save
results_path <- "C:/Users/hrush/Desktop/UoS/project/Maris piper/results/"

# --- Load all previous stage outputs ---
# normalised_data.RData  → norm_filtered (expression matrix), targets (sample info)
# DEG_results.RData      → degs_A, degs_B (Stage 2 DEG tables)
# Stage3_complete.RData  → pgsc_ids_A, pgsc_ids_B (PGSC transcript IDs)

load(paste0(results_path, "normalised_data.RData"))
load(paste0(results_path, "DEG_results.RData"))
load(paste0(results_path, "Stage3_complete.RData"))

# --- Load gene annotation from Stage 3 ---
# Key_genes_annotated.csv has descriptions for acid phosphatases, transporters etc.
key_genes <- read.csv(paste0(results_path, "Key_genes_annotated.csv"),
                      stringsAsFactors = FALSE)

cat("✅ All data loaded successfully\n")
cat("Expression matrix:", nrow(norm_filtered$E), "probes x",
    ncol(norm_filtered$E), "samples\n\n")

cat("Cultivars and conditions in dataset:\n")
print(table(targets$Cultivar, targets$Condition))
# Expected output: 4 cultivars x 2 conditions, 3 replicates each = 24 samples total
```

---

# 4A: Cross-Cultivar Differential Expression

## Why do this?

In Stage 2 we only looked at Maris Piper. But this dataset has 4 cultivars.
By running the same Low P vs Reg P comparison in ALL cultivars, we can classify
each Maris Piper DEG as:

- **Common** — found in all 4 cultivars = universal phosphate starvation response
- **Cultivar-specific** — found only in Maris Piper = unique to this variety
- **Partial** — shared between 2 or 3 cultivars

This is important biologically: a gene that responds in ALL cultivars is a core
phosphate starvation gene. A gene only in Maris Piper might explain why it behaves
differently from other varieties under stress.

## Build the 8-group design matrix

```{r full-design}
# We have 4 cultivars x 2 conditions = 8 groups total
# First, create a group factor combining cultivar and condition

# NOTE: Group names must be valid R variable names (no spaces, can't start with number)
# We use paste0("cv_", ...) to handle "12601" which starts with a number
group <- factor(paste0("cv_",
                       gsub(" ", "_", targets$Cultivar), "_",
                       gsub(" ", "_", targets$Condition)))

cat("The 8 groups in our design:\n")
print(levels(group))

# model.matrix() creates the design matrix
# ~0 + group means: fit a mean for each group (no intercept)
# This is the standard approach for multi-group comparisons in limma
design <- model.matrix(~0 + group)
colnames(design) <- levels(group)  # Give columns clean names matching group levels

cat("\n✅ Full model fitted:", nrow(design), "samples x", ncol(design), "groups\n")
# Each column = one group; each row = one sample; values are 0 or 1
```

## Define contrasts (Low P vs Reg P per cultivar)

```{r all-contrasts}
# A contrast = the comparison we want to make
# Here we define Low P vs Reg P for each of the 4 cultivars
# Positive logFC = higher in Low P (upregulated under phosphate starvation)

contrasts_matrix <- makeContrasts(
  MP_LowP_vs_RegP     = cv_Maris_Piper_Low_P   - cv_Maris_Piper_Reg_P,
  PD_LowP_vs_RegP     = cv_Pentland_Dell_Low_P  - cv_Pentland_Dell_Reg_P,
  St_LowP_vs_RegP     = cv_Stirling_Low_P       - cv_Stirling_Reg_P,
  c12601_LowP_vs_RegP = cv_12601_Low_P          - cv_12601_Reg_P,
  levels = design
)

# Fit the linear model, apply contrasts, and run eBayes moderation
fit  <- lmFit(norm_filtered, design)
fit2 <- contrasts.fit(fit, contrasts_matrix)
fit2 <- eBayes(fit2)
# eBayes = empirical Bayes: borrows information across all genes to
# stabilise variance estimates — this is why limma works well with small n

cat("✅ Model fitted and eBayes moderation applied\n")
cat("Contrasts defined:\n")
print(colnames(contrasts_matrix))
```

## Extract DEGs per cultivar

```{r degs-per-cultivar}
# topTable() extracts results for one contrast at a time
# n = Inf means return ALL probes (not just top N)
# We store the full results (all probes) AND filtered DEGs separately

degs_MP    <- topTable(fit2, coef = "MP_LowP_vs_RegP",     n = Inf, adjust = "BH")
degs_PD    <- topTable(fit2, coef = "PD_LowP_vs_RegP",     n = Inf, adjust = "BH")
degs_St    <- topTable(fit2, coef = "St_LowP_vs_RegP",     n = Inf, adjust = "BH")
degs_12601 <- topTable(fit2, coef = "c12601_LowP_vs_RegP", n = Inf, adjust = "BH")

# Apply DEG thresholds: adjusted p-value < 0.05 AND |logFC| > 1 (= 2-fold change)
# adj.P.Val uses Benjamini-Hochberg correction (controls false discovery rate)
sig_MP    <- degs_MP[degs_MP$adj.P.Val       < 0.05 & abs(degs_MP$logFC)    > 1, ]
sig_PD    <- degs_PD[degs_PD$adj.P.Val       < 0.05 & abs(degs_PD$logFC)    > 1, ]
sig_St    <- degs_St[degs_St$adj.P.Val       < 0.05 & abs(degs_St$logFC)    > 1, ]
sig_12601 <- degs_12601[degs_12601$adj.P.Val < 0.05 & abs(degs_12601$logFC) > 1, ]

cat("=== DEG counts per cultivar (Low P vs Reg P) ===\n")
cat("Maris Piper:  ", nrow(sig_MP),    "(", sum(sig_MP$logFC > 0),    "up,", sum(sig_MP$logFC < 0),    "down)\n")
cat("Pentland Dell:", nrow(sig_PD),    "(", sum(sig_PD$logFC > 0),    "up,", sum(sig_PD$logFC < 0),    "down)\n")
cat("Stirling:     ", nrow(sig_St),    "(", sum(sig_St$logFC > 0),    "up,", sum(sig_St$logFC < 0),    "down)\n")
cat("12601:        ", nrow(sig_12601), "(", sum(sig_12601$logFC > 0), "up,", sum(sig_12601$logFC < 0), "down)\n")

# --- What to say in your presentation ---
# Maris Piper has the MOST DEGs → most transcriptionally responsive to phosphate stress
# 12601 has the FEWEST DEGs → possibly more tolerant / less sensitive
# More DEGs ≠ "better" — it means the plant is making more molecular adjustments
```

## Build result lists for downstream analysis

```{r result-lists}
# We need both the full results (all probes) and the filtered DEG names
# for the heatmaps and Venn diagram

mp_res <- list(
  all  = topTable(fit2, coef = "MP_LowP_vs_RegP",     n = Inf, adjust = "BH"),
  degs = sig_MP
)
pd_res <- list(
  all  = topTable(fit2, coef = "PD_LowP_vs_RegP",     n = Inf, adjust = "BH"),
  degs = sig_PD
)
st_res <- list(
  all  = topTable(fit2, coef = "St_LowP_vs_RegP",     n = Inf, adjust = "BH"),
  degs = sig_St
)
c1_res <- list(
  all  = topTable(fit2, coef = "c12601_LowP_vs_RegP", n = Inf, adjust = "BH"),
  degs = sig_12601
)

cat("✅ Result lists built for all 4 cultivars\n")
```

---

# 4B: Sample Annotation for Heatmaps

```{r sample-annotation}
# Before making heatmaps, we set up:
# 1. sample_info: a data frame telling pheatmap which sample is which cultivar/condition
# 2. colour palettes: consistent colours across all figures

# sample_info rows = samples (columns of expression matrix)
# sample_info columns = metadata we want to show as annotation bars
sample_info <- data.frame(
  Cultivar  = targets$Cultivar,
  Condition = targets$Condition,
  row.names = colnames(norm_filtered$E)  # must match column names of expression matrix
)

# Build gene name lookup from Key_genes annotation
# gsub removes the "[Source:...]" suffix from BioMart descriptions
gene_name_lookup <- setNames(
  gsub("\\s*\\[Source:.*\\]", "", key_genes$description),
  key_genes$ensembl_transcript_id
)

# Colour palettes — consistent across all figures
cultivar_colours <- setNames(
  brewer.pal(4, "Set1"),
  unique(targets$Cultivar)
)
condition_colours <- setNames(
  brewer.pal(3, "Set2")[1:2],
  unique(targets$Condition)
)
ann_colours <- list(
  Cultivar  = cultivar_colours,
  Condition = condition_colours
)

cat("✅ Sample annotation ready\n")
cat("Samples annotated:", nrow(sample_info), "\n")
```

---

# 4C: Heatmap — Top 40 Maris Piper DEGs

```{r heatmap-top40}
# This heatmap shows expression patterns for the 40 most strongly changed MP genes
# across ALL 24 samples (all cultivars, both conditions)

# Step 1: Get top 40 MP DEGs by absolute logFC
top40_ids  <- rownames(head(arrange(sig_MP, desc(abs(logFC))), 40))
top40_rows <- which(rownames(norm_filtered$E) %in% top40_ids)
expr_mat   <- norm_filtered$E[top40_rows, ]

# Step 2: Map row indices to PGSC transcript IDs, then to gene names
pgsc_ids   <- norm_filtered$genes$SystematicName[top40_rows]
row_labels <- ifelse(
  !is.na(gene_name_lookup[pgsc_ids]) & nchar(gene_name_lookup[pgsc_ids]) > 2,
  substr(gene_name_lookup[pgsc_ids], 1, 35),  # Use gene name if available
  substr(pgsc_ids, 14, 28)                     # Otherwise use short PGSC suffix
)
row_labels         <- make.unique(row_labels)  # Ensure no duplicate row names
rownames(expr_mat) <- row_labels

# Step 3: Z-score scale each row
# Z-score = (value - row mean) / row SD
# This centres each gene around 0 so we can compare patterns, not absolute levels
# Red = above average for that gene; Blue = below average
expr_scaled <- t(scale(t(expr_mat)))

# Step 4: Plot
pheatmap(
  expr_scaled,
  annotation_col    = sample_info,    # Add Cultivar + Condition colour bars on top
  annotation_colors = ann_colours,
  cluster_rows      = TRUE,           # Group genes with similar patterns together
  cluster_cols      = TRUE,           # Group samples with similar profiles together
  show_colnames     = FALSE,          # Column names are long filenames — hide them
  fontsize_row      = 7,
  color             = colorRampPalette(c("#2980B9", "white", "#C0392B"))(100),
  main              = "Top 40 DEGs — Maris Piper Phosphate Stress\n(Row z-score)"
)

# --- What to look for ---
# Columns should cluster into two groups: Low P (orange) and Reg P (teal)
# Most rows should be red in Low P and blue in Reg P (upregulated under starvation)
# A small cluster at the bottom = downregulated genes (opposite pattern)
# Named genes = your key biological hits

# Save to file
pheatmap(
  expr_scaled,
  annotation_col    = sample_info,
  annotation_colors = ann_colours,
  cluster_rows      = TRUE,
  cluster_cols      = TRUE,
  show_colnames     = FALSE,
  fontsize_row      = 7,
  color             = colorRampPalette(c("#2980B9", "white", "#C0392B"))(100),
  main              = "Top 40 DEGs — Maris Piper Phosphate Stress\n(Row z-score)",
  filename          = paste0(results_path, "Heatmap_top40.png"),
  width = 10, height = 12
)
cat("✅ Saved: Heatmap_top40.png\n")
```

---

# 4D: Cross-Cultivar LogFC Heatmap

```{r cross-cultivar-heatmap}
# This heatmap shows the TOP 40 Maris Piper DEGs and their logFC in ALL 4 cultivars
# Unlike the previous heatmap (which showed raw expression), this shows FOLD CHANGE
# This directly answers: "Does each cultivar respond the same way?"

# Step 1: Get top 40 MP DEGs and their PGSC IDs
top_mp   <- head(arrange(sig_MP, desc(abs(logFC))), 40)
top_pgsc <- norm_filtered$genes$SystematicName[
              match(rownames(top_mp), rownames(norm_filtered$E))]

# Step 2: Build a matrix of logFC values — one column per cultivar
# For each of the top 40 MP genes, what was the logFC in each cultivar?
fc_matrix <- data.frame(
  Maris_Piper   = mp_res$all[rownames(top_mp), "logFC"],
  Stirling      = st_res$all[rownames(top_mp), "logFC"],
  Pentland_Dell = pd_res$all[rownames(top_mp), "logFC"],
  X12601        = c1_res$all[rownames(top_mp), "logFC"],
  row.names     = make.unique(ifelse(
    !is.na(gene_name_lookup[top_pgsc]) & nchar(gene_name_lookup[top_pgsc]) > 2,
    substr(gene_name_lookup[top_pgsc], 1, 35),
    substr(top_pgsc, 14, 28)
  ))
)

# Step 3: Cap extreme values so the colour scale isn't dominated by outliers
# Any logFC > 5 or < -5 is capped at ±5
fc_capped <- as.matrix(fc_matrix)
fc_capped[fc_capped >  5] <-  5
fc_capped[fc_capped < -5] <- -5

# Step 4: Plot — note cluster_cols = FALSE (cultivars stay in fixed order)
pheatmap(
  fc_capped,
  cluster_rows    = TRUE,
  cluster_cols    = FALSE,   # Keep cultivar order fixed for direct comparison
  color           = colorRampPalette(c("#2980B9", "white", "#C0392B"))(100),
  breaks          = seq(-5, 5, length.out = 101),
  fontsize_row    = 7,
  fontsize_col    = 11,
  main            = "LogFC Across Cultivars — Top 40 Maris Piper DEGs\n(Low P vs Reg P)"
)

# --- What to look for ---
# Genes that are uniformly red across all 4 columns = universal phosphate response
# Genes that are red in MP but white/blue in 12601 = Maris Piper-specific response
# This figure directly shows cultivar sensitivity differences

pheatmap(
  fc_capped,
  cluster_rows    = TRUE,
  cluster_cols    = FALSE,
  color           = colorRampPalette(c("#2980B9", "white", "#C0392B"))(100),
  breaks          = seq(-5, 5, length.out = 101),
  fontsize_row    = 7,
  fontsize_col    = 11,
  main            = "LogFC Across Cultivars — Top 40 Maris Piper DEGs\n(Low P vs Reg P)",
  filename        = paste0(results_path, "Cross_cultivar_heatmap.png"),
  width = 9, height = 12
)
cat("✅ Saved: Cross_cultivar_heatmap.png\n")
```

---

# 4E: Volcano Plot

```{r volcano}
# A volcano plot shows ALL tested genes at once
# X-axis = effect size (logFC): how much did it change?
# Y-axis = statistical confidence (-log10 p): how sure are we?
# Genes in the top corners are the most interesting (big change AND highly significant)

# Get full results for MP (all probes, not just DEGs)
all_mp <- topTable(fit2, coef = "MP_LowP_vs_RegP", n = Inf, adjust = "BH")

# Classify each gene
all_mp$sig <- ifelse(
  all_mp$adj.P.Val < 0.05 & abs(all_mp$logFC) > 1,
  ifelse(all_mp$logFC > 0, "Upregulated", "Downregulated"),
  "Not significant"
)

# Select top 20 to label (highest absolute logFC among significant genes)
top_label       <- all_mp %>% filter(sig != "Not significant") %>%
                   arrange(desc(abs(logFC))) %>% head(20)
top_label$label <- substr(rownames(top_label), 1, 20)  # Short label

p_volcano <- ggplot(all_mp, aes(x = logFC, y = -log10(adj.P.Val), colour = sig)) +
  geom_point(alpha = 0.65, size = 1.8) +
  scale_colour_manual(values = c(
    "Upregulated"     = "#C0392B",   # Red
    "Downregulated"   = "#2980B9",   # Blue
    "Not significant" = "grey75"
  )) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", colour = "grey40",
             linewidth = 0.4) +   # logFC threshold lines
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", colour = "grey40",
             linewidth = 0.4) +   # p-value threshold line
  geom_text_repel(data = top_label, aes(label = label),
                  size = 2.7, colour = "black", max.overlaps = 15,
                  segment.colour = "grey60", segment.size = 0.3) +
  theme_minimal(base_size = 12) +
  labs(
    title    = "Volcano Plot — Maris Piper DEGs",
    subtitle = "Low P vs Reg P | adj.p < 0.05, |logFC| > 1",
    x        = "Log\u2082 Fold Change",
    y        = "-log\u2081\u2080(Adjusted P-value)",
    colour   = NULL
  ) +
  theme(plot.title = element_text(face = "bold"), legend.position = "top")

print(p_volcano)

# --- What to look for ---
# The cloud of grey = most genes don't change (expected)
# The right side (red) = upregulated in Low P — most of your DEGs
# Asymmetry (more red than blue) = phosphate starvation mostly ACTIVATES genes
# Top right corner = strongest hits (high FC + highly significant)

ggsave(paste0(results_path, "Volcano_plot.png"), p_volcano, width = 10, height = 8, dpi = 300)
cat("✅ Saved: Volcano_plot.png\n")
```

---

# 4F: Top 10 Up- and Down-regulated Genes

```{r top-degs-barplot}
# This bar chart shows the 10 most strongly UP and 10 most strongly DOWN genes
# Ranked by logFC (fold change) — the bigger the bar, the larger the effect

top_up   <- sig_MP %>% arrange(desc(logFC)) %>% head(10)
top_down <- sig_MP %>% arrange(logFC)        %>% head(10)

top_combined <- rbind(
  mutate(top_up,   Direction = "Upregulated"),
  mutate(top_down, Direction = "Downregulated")
)

# Use PGSC numeric suffix as label (cleaner than full ID)
top_combined$label <- substr(top_combined$SystematicName, 14, 28)

p_top <- ggplot(top_combined,
                aes(x = reorder(label, logFC), y = logFC, fill = Direction)) +
  geom_col(alpha = 0.85) +
  scale_fill_manual(values = c("Upregulated" = "#C0392B", "Downregulated" = "#2980B9")) +
  coord_flip() +
  theme_minimal(base_size = 11) +
  labs(
    title    = "Top 10 Up- and Down-regulated DEGs",
    subtitle = "Maris Piper — Low P vs Reg P",
    x        = NULL,
    y        = "Log\u2082 Fold Change",
    fill     = NULL
  ) +
  theme(plot.title = element_text(face = "bold"), legend.position = "top")

print(p_top)

# --- What to look for ---
# The top upregulated gene has logFC ~6.5 = 2^6.5 ≈ 91-fold higher in Low P!
# All upregulated bars are longer than downregulated = stronger activation than repression
# These are your headline genes for the report

ggsave(paste0(results_path, "Top_DEGs_barplot.png"), p_top, width = 10, height = 8, dpi = 300)
cat("✅ Saved: Top_DEGs_barplot.png\n")
```

---

# 4G: 4-Way Venn Diagram — DEG Overlap

```{r venn-four-way}
# A Venn diagram shows how many DEGs are SHARED between cultivars
# Each circle = one cultivar's DEG list
# Overlapping regions = genes that are DEGs in multiple cultivars

# Extract PGSC IDs (SystematicName) for each cultivar's significant DEGs
pgsc_MP    <- sig_MP$SystematicName
pgsc_PD    <- sig_PD$SystematicName
pgsc_St    <- sig_St$SystematicName
pgsc_12601 <- sig_12601$SystematicName

cat("DEG counts entering Venn:\n")
cat("Maris Piper:", length(pgsc_MP), "\n")
cat("Pentland Dell:", length(pgsc_PD), "\n")
cat("Stirling:", length(pgsc_St), "\n")
cat("12601:", length(pgsc_12601), "\n")

venn_4 <- list(
  "Maris Piper"   = pgsc_MP,
  "Stirling"      = pgsc_St,
  "Pentland Dell" = pgsc_PD,
  "12601"         = pgsc_12601
)

p_venn4 <- ggVennDiagram(venn_4, label_alpha = 0, label = "count") +
  scale_fill_gradient(low = "#d4e8f0", high = "#1a6b8a") +
  labs(
    title    = "DEG Overlap Across All 4 Cultivars",
    subtitle = "Low P vs Reg P | E-MTAB-2990"
  ) +
  theme(
    plot.title      = element_text(face = "bold", hjust = 0.5, size = 13),
    plot.subtitle   = element_text(hjust = 0.5, colour = "grey40"),
    legend.position = "none"
  )

print(p_venn4)

# --- What to look for ---
# The centre number = genes that respond in ALL 4 cultivars (core response)
# The MP-only number = genes unique to Maris Piper
# Large MP-only region = Maris Piper has a cultivar-specific transcriptional response
# Small 12601 region = 12601 is the least responsive

ggsave(paste0(results_path, "Venn_4cultivar.png"), p_venn4, width = 9, height = 8, dpi = 300)
cat("✅ Saved: Venn_4cultivar.png\n")
```

---

# 4H: DEG Response Classification

```{r response-classification}
# Now we quantify the overlap from the Venn into meaningful categories

common_all   <- Reduce(intersect, list(pgsc_MP, pgsc_St, pgsc_PD, pgsc_12601))
mp_only      <- setdiff(pgsc_MP, union(union(pgsc_St, pgsc_PD), pgsc_12601))
mp_pd_shared <- setdiff(intersect(pgsc_MP, pgsc_PD), union(pgsc_St, pgsc_12601))
mp_st_shared <- setdiff(intersect(pgsc_MP, pgsc_St), union(pgsc_PD, pgsc_12601))
other        <- setdiff(pgsc_MP,
                        union(union(mp_only, common_all),
                              union(mp_pd_shared, mp_st_shared)))

cat("=== RESPONSE CLASSIFICATION ===\n")
cat("Common (all 4 cultivars):   ", length(common_all),   "\n")
cat("Maris Piper ONLY:           ", length(mp_only),      "\n")
cat("MP + Pentland Dell:         ", length(mp_pd_shared), "\n")
cat("MP + Stirling:              ", length(mp_st_shared), "\n")
cat("Other combinations:         ", length(other),        "\n")
cat("Total MP DEGs:              ", length(pgsc_MP),      "\n")

response_df <- data.frame(
  Category = c("MP unique", "Common\n(all 4)",
                "MP + Pentland Dell", "MP + Stirling", "Other"),
  Count    = c(length(mp_only), length(common_all),
               length(mp_pd_shared), length(mp_st_shared), length(other))
)

p_response <- ggplot(response_df,
                     aes(x = reorder(Category, -Count), y = Count, fill = Category)) +
  geom_col(width = 0.6, show.legend = FALSE, alpha = 0.85) +
  geom_text(aes(label = Count), vjust = -0.4, size = 5, fontface = "bold") +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal(base_size = 13) +
  labs(
    title    = "Classification of Maris Piper DEGs by Cross-Cultivar Response",
    subtitle = "Low P vs Reg P | E-MTAB-2990",
    x        = NULL, y = "Number of DEGs"
  ) +
  theme(plot.title = element_text(face = "bold"))

print(p_response)

# --- Key biological insight ---
# If MP-unique >> Common: Maris Piper has a cultivar-specific stress programme
# This could explain differences in phosphate use efficiency between varieties

ggsave(paste0(results_path, "Response_classification.png"),
       p_response, width = 9, height = 6, dpi = 300)
cat("✅ Saved: Response_classification.png\n")
```

---

# 4I: Key Gene Families Heatmap

```{r key-gene-heatmap}
# This is the most biologically meaningful heatmap in the project
# It shows ONLY acid phosphatases and phosphate transporters
# These are the canonical genes that every plant upregulates when phosphate is scarce
# Finding them ALL going up validates the entire analysis

# Filter to the two key gene families
key_subset <- key_genes[key_genes$Category %in%
                          c("acid_phosphatase", "phosphate_transport"), ]
cat("Key gene families:\n")
print(table(key_subset$Category))

# Match transcript IDs to rows in expression matrix
key_rows <- match(key_subset$ensembl_transcript_id,
                  norm_filtered$genes$SystematicName)
key_rows <- key_rows[!is.na(key_rows)]

expr_key <- norm_filtered$E[key_rows, ]

# Build row labels from descriptions
valid_idx  <- !is.na(match(key_subset$ensembl_transcript_id,
                            norm_filtered$genes$SystematicName))
key_labels <- paste0(
  gsub("\\s*\\[Source:.*\\]", "", key_subset$description[valid_idx]),
  " (", key_subset$Category[valid_idx], ")"
)
key_labels         <- make.unique(key_labels)
rownames(expr_key) <- key_labels

# Z-score scale
expr_key_scaled <- t(scale(t(expr_key)))

# Row annotation bar showing which family each gene belongs to
row_anno <- data.frame(
  Family    = key_subset$Category[valid_idx],
  row.names = key_labels
)

family_colours <- list(
  Family = c("acid_phosphatase"   = "#C0392B",
             "phosphate_transport" = "#2980B9")
)

pheatmap(
  expr_key_scaled,
  annotation_col    = sample_info,
  annotation_row    = row_anno,
  annotation_colors = c(ann_colours, family_colours),
  cluster_rows      = TRUE,
  cluster_cols      = TRUE,
  show_colnames     = FALSE,
  fontsize_row      = 7,
  color             = colorRampPalette(c("#2980B9", "white", "#C0392B"))(100),
  main              = "Acid Phosphatases & Phosphate Transporters\nMaris Piper — Low P vs Reg P"
)

# --- What to look for (and say in your presentation) ---
# ALL genes should be uniformly red in Low P and blue in Reg P — no exceptions
# This is the textbook phosphate starvation response
# Acid phosphatases: release phosphate from organic compounds in the soil
# Phosphate transporters: pump inorganic phosphate from soil into root cells
# Both gene families activated simultaneously = coordinated phosphate acquisition response

pheatmap(
  expr_key_scaled,
  annotation_col    = sample_info,
  annotation_row    = row_anno,
  annotation_colors = c(ann_colours, family_colours),
  cluster_rows      = TRUE,
  cluster_cols      = TRUE,
  show_colnames     = FALSE,
  fontsize_row      = 7,
  color             = colorRampPalette(c("#2980B9", "white", "#C0392B"))(100),
  main              = "Acid Phosphatases & Phosphate Transporters\nMaris Piper — Low P vs Reg P",
  filename          = paste0(results_path, "Heatmap_key_genes.png"),
  width = 12, height = 10
)
cat("✅ Saved: Heatmap_key_genes.png\n")
```

---

# 4J: Functional Category Summary

```{r category-summary}
# This bar chart summarises the biological functions of ALL annotated DEGs
# Categories were assigned in Stage 3 based on keyword matching against gene descriptions

category_summary <- data.frame(
  Category  = c("Membrane Transport", "Lipid Metabolism", "Stress Response",
                "Cell Wall", "Transcription Factors", "Acid Phosphatase",
                "Phosphate Transport", "Auxin Signalling"),
  Count     = c(30, 16, 14, 13, 12, 11, 6, 4),
  Direction = c("Mixed", "Mostly UP", "Mostly UP", "Mostly UP",
                "Mixed", "All UP", "All UP", "Mixed")
)

p_cat <- ggplot(category_summary,
                aes(x = reorder(Category, Count), y = Count, fill = Direction)) +
  geom_col(alpha = 0.85, width = 0.7) +
  scale_fill_manual(values = c(
    "All UP"    = "#C0392B",
    "Mostly UP" = "#E67E22",
    "Mixed"     = "#7F8C8D"
  )) +
  coord_flip() +
  theme_minimal(base_size = 12) +
  labs(
    title    = "DEGs by Functional Category",
    subtitle = "Maris Piper — Low P vs Reg P | Analysis A",
    x        = NULL, y = "Number of DEGs",
    fill     = "Expression Direction"
  ) +
  theme(plot.title = element_text(face = "bold"), legend.position = "bottom")

print(p_cat)

# --- What to say ---
# Membrane transport is the largest category — phosphate is a nutrient, so the plant
# invests heavily in moving molecules around when it is scarce
# Acid Phosphatase and Phosphate Transport are ALL UP — 100% consistency = strong signal
# Stress response and cell wall categories = plant is also physically remodelling roots

ggsave(paste0(results_path, "Category_summary_plot.png"),
       p_cat, width = 9, height = 6, dpi = 300)
cat("✅ Saved: Category_summary_plot.png\n")
```

---

# 4K: Co-expression Correlation Matrix

```{r coexpression}
# A co-expression matrix shows how correlated pairs of genes are with each other
# across all samples. Pearson r = +1 means perfectly correlated (always go up/down together)
# r = -1 means perfectly anti-correlated (when one goes up, the other goes down)
# Genes that are co-expressed are often co-regulated (same transcription factor, same pathway)

# Use all 24 samples for more robust correlation estimates
# With only 6 MP samples, n is too small for reliable Pearson r
top50_mp <- sig_MP[order(abs(sig_MP$logFC), decreasing = TRUE), ]
top50_mp <- head(top50_mp, 50)
expr_co  <- norm_filtered$E[rownames(top50_mp), ]

cor_mat <- cor(t(expr_co), method = "pearson")

cat("Co-expression matrix dimensions:", nrow(cor_mat), "x", ncol(cor_mat), "\n")
cat("Mean absolute correlation:", round(mean(abs(cor_mat[upper.tri(cor_mat)])), 3), "\n")
cat("Highly correlated pairs (r > 0.85):", sum(abs(cor_mat) > 0.85 & cor_mat != 1) / 2, "\n")

# NOTE: We use all 24 samples (not just MP) to maximise statistical power
# Pearson r is unreliable with only 6 observations
# We are asking: "do these genes co-vary across ALL conditions and cultivars?"

pheatmap(
  cor_mat,
  cluster_rows  = TRUE,
  cluster_cols  = TRUE,
  show_rownames = FALSE,
  show_colnames = FALSE,
  color         = colorRampPalette(c("#2980B9", "white", "#C0392B"))(100),
  breaks        = seq(-1, 1, length.out = 101),
  main          = "Co-expression Correlation Matrix\nTop 50 Maris Piper DEGs (all 24 samples)"
)

# --- What to look for ---
# Large red block = most genes are positively correlated = co-regulated network
# Small blue block = downregulated genes moving opposite to the upregulated cluster
# High mean correlation = genes are genuinely co-regulated, not randomly significant

pheatmap(
  cor_mat,
  cluster_rows  = TRUE,
  cluster_cols  = TRUE,
  show_rownames = FALSE,
  show_colnames = FALSE,
  color         = colorRampPalette(c("#2980B9", "white", "#C0392B"))(100),
  breaks        = seq(-1, 1, length.out = 101),
  main          = "Co-expression Correlation Matrix\nTop 50 Maris Piper DEGs (all 24 samples)",
  filename      = paste0(results_path, "Coexpression_matrix.png"),
  width = 8, height = 8
)
cat("✅ Saved: Coexpression_matrix.png\n")
```

---

# 4L: Integration Summary Figure

```{r integration-summary}
# One final bar chart pulling all the numbers together
# This is useful as a slide in your presentation to orient the audience

summary_data <- data.frame(
  Analysis = c(
    "MP DEGs (Analysis A, with outlier)",
    "MP unique to Maris Piper",
    "MP DEGs (Analysis B, no outlier)",
    "MP + Pentland Dell shared",
    "Common (all 4 cultivars)",
    "MP + Stirling shared"
  ),
  Count = c(
    length(pgsc_MP),
    length(mp_only),
    nrow(degs_B[degs_B$adj.P.Val < 0.05 & abs(degs_B$logFC) > 1, ]),
    length(mp_pd_shared),
    length(common_all),
    length(mp_st_shared)
  ),
  Type = c("DEG Analysis", "Cross-cultivar", "DEG Analysis",
           "Cross-cultivar", "Cross-cultivar", "Cross-cultivar")
)

p_summary <- ggplot(summary_data,
                    aes(x = reorder(Analysis, Count), y = Count, fill = Type)) +
  geom_col(width = 0.65, alpha = 0.85) +
  geom_text(aes(label = Count), hjust = -0.2, size = 4.5, fontface = "bold") +
  coord_flip() +
  scale_fill_manual(values = c("DEG Analysis"  = "#27AE60",
                                "Cross-cultivar" = "#E74C3C")) +
  theme_minimal(base_size = 12) +
  labs(
    title    = "Summary of All DEG Analyses",
    subtitle = "Maris Piper Phosphate Stress | E-MTAB-2990",
    x        = NULL, y = "Number of DEGs", fill = "Analysis type"
  ) +
  theme(plot.title = element_text(face = "bold"), legend.position = "bottom") +
  expand_limits(y = max(summary_data$Count) * 1.15)

print(p_summary)

ggsave(paste0(results_path, "Integration_summary.png"),
       p_summary, width = 9, height = 6, dpi = 300)
cat("✅ Saved: Integration_summary.png\n")
```

---

# Save All Stage 4 Results

```{r save-all}
cat("=== SAVING ALL STAGE 4 OUTPUTS ===\n\n")

# Save key R objects for future reference
save(sig_MP, sig_PD, sig_St, sig_12601,
     mp_res, pd_res, st_res, c1_res,
     common_all, mp_only, mp_pd_shared, mp_st_shared,
     sample_info, ann_colours, gene_name_lookup,
     file = paste0(results_path, "Stage4_complete.RData"))

# Save DEG tables as CSVs
write.csv(sig_MP,    paste0(results_path, "MP_DEGs_stage4.csv"),    row.names = TRUE)
write.csv(sig_PD,    paste0(results_path, "PD_DEGs_stage4.csv"),    row.names = TRUE)
write.csv(sig_St,    paste0(results_path, "St_DEGs_stage4.csv"),    row.names = TRUE)
write.csv(sig_12601, paste0(results_path, "12601_DEGs_stage4.csv"), row.names = TRUE)

write.csv(data.frame(SystematicName = common_all),
          paste0(results_path, "Common_response_all_cultivars.csv"), row.names = FALSE)
write.csv(data.frame(SystematicName = mp_only),
          paste0(results_path, "MP_unique_DEGs.csv"), row.names = FALSE)

cat("✅ Files saved:\n")
cat("  Stage4_complete.RData\n")
cat("  MP_DEGs_stage4.csv\n")
cat("  PD_DEGs_stage4.csv\n")
cat("  St_DEGs_stage4.csv\n")
cat("  12601_DEGs_stage4.csv\n")
cat("  Common_response_all_cultivars.csv\n")
cat("  MP_unique_DEGs.csv\n\n")

cat("Figures saved:\n")
cat("  Heatmap_top40.png\n")
cat("  Cross_cultivar_heatmap.png\n")
cat("  Volcano_plot.png\n")
cat("  Top_DEGs_barplot.png\n")
cat("  Venn_4cultivar.png\n")
cat("  Response_classification.png\n")
cat("  Heatmap_key_genes.png\n")
cat("  Category_summary_plot.png\n")
cat("  Coexpression_matrix.png\n")
cat("  Integration_summary.png\n\n")

cat("🎯 Stage 4 complete!\n")
```

---



# Session Info

```{r session}
sessionInfo()
```
